\chapter{Modern \cpp{} for \maya{}}
\label{ch:cpp}

This chapter is a self-contained course in the slice of \cpp{} that
\maya{} uses. It assumes no \cpp{} background. If you have written
\cpp{}11 or \cpp{}14 it will still be useful, because \maya{} leans
on features that arrived in \cpp{}20, \cpp{}23, and \cpp{}26; some of
the patterns will be new even to experienced \cpp{} programmers.

Read it linearly the first time. The sections build on each other:
later ones (templates, concepts, variants) lean on earlier ones
(values, references, classes). When you finish, you will be able to
read every line of \maya{} source code.

\section{What a \cpp{} program is}
\label{sec:cpp:program}

A \cpp{} program is one or more \textbf{source files} compiled
together into an \textbf{executable}. The smallest possible one fits
on one line, although by convention we spread it out:

\begin{lstlisting}
#include <print>

int main() {
    std::println("Hello, world");
}
\end{lstlisting}

Four things are happening, and each deserves a paragraph.

\paragraph{The preprocessor runs first.}
Before the compiler proper looks at your code, a small program called
the \textbf{preprocessor} does textual substitution. Lines beginning
with \code{\#} are its instructions. \code{\#include <print>}
literally means ``find the file named \file{print} in the standard
library's directories and paste its contents here''. After the
preprocessor finishes, your source file has grown to several
thousand lines, all the standard-library declarations now visible.

\paragraph{The compiler reads what is left.}
The expanded text is parsed, type-checked, and translated into
\textbf{object code} --- machine instructions in a not-yet-runnable
format. Each source file becomes one \emph{object file}, also called
a \textbf{translation unit}. A small project has one; \maya{} has
several hundred.

\paragraph{The linker stitches object files together.}
The \textbf{linker} takes all object files plus any libraries you
depend on and produces a single executable. Its main job is
resolving \textbf{symbols}: when one file uses a function defined in
another file, the linker matches the use to the definition.

\paragraph{The operating system runs the result.}
When you type \code{./hello}, the OS loads the executable into
memory and jumps to a specific named function: \code{main}. The
integer \code{main} returns is the process's \emph{exit status}
(zero = success). The whole program ends when \code{main} returns.

\begin{idea}
The preprocessor pastes text. The compiler turns text into object
files. The linker joins object files into an executable. The OS runs
the executable from \code{main}.
\end{idea}

\subsection{Headers, source files, and the one-definition rule}

A \textbf{header file} (suffix \file{.hpp} or \file{.h}) is a file
you \code{\#include} into other files. It typically contains
\emph{declarations} --- statements about what exists, but not how it
works. A \textbf{source file} (suffix \file{.cpp}) typically contains
\emph{definitions} --- the actual function bodies and data.

\begin{lstlisting}
// math.hpp --- declaration
int add(int a, int b);
\end{lstlisting}

\begin{lstlisting}
// math.cpp --- definition
#include "math.hpp"
int add(int a, int b) { return a + b; }
\end{lstlisting}

\begin{lstlisting}
// main.cpp --- user
#include "math.hpp"
int main() { return add(2, 3); }
\end{lstlisting}

Every use of \code{add} in any source file sees the same declaration
(from \file{math.hpp}), and there is exactly one definition in the
whole program (in \file{math.cpp}). That is the \textbf{One Definition
Rule}, ODR: every entity must have exactly one definition across all
translation units linked together.

\maya{} is a \emph{header-mostly} library. Almost all its code lives
in headers, with most functions and classes defined \code{inline} or
as templates. This is unusual and has trade-offs (compile times can
be long), but it lets the compiler aggressively inline and constant-
fold across the library boundary. We discuss the trade-off in
Chapter~\ref{ch:build}.

\subsection{\code{\#pragma once} and include guards}

If two headers include the same third header, the preprocessor would
paste it twice, producing duplicate declarations. The traditional
fix is an \textbf{include guard}:

\begin{lstlisting}
// math.hpp
#ifndef MATH_HPP
#define MATH_HPP
int add(int a, int b);
#endif
\end{lstlisting}

The second inclusion finds \code{MATH\_HPP} already defined and skips
the contents. Modern code uses the shorter and equivalent:

\begin{lstlisting}
// math.hpp
#pragma once
int add(int a, int b);
\end{lstlisting}

Every file in \maya{} starts with \code{\#pragma once}.

\subsection{How to actually compile and run a \cpp{} file}

You will get more out of the rest of this chapter if you can try
things as you read. The minimum tool set is a compiler and a
terminal. On Linux or WSL:

\begin{lstlisting}[style=shell]
sudo apt install g++-15        # or your distro's equivalent
echo 'int main() {}' > t.cpp
g++-15 -std=c++26 t.cpp -o t   # compile
./t                            # run
\end{lstlisting}

On macOS install a recent Homebrew \code{llvm} and use \code{clang++}
with \code{-std=c++26}. On Windows the easiest path is WSL with a
current Ubuntu.

The useful compiler flags for learning:

\begin{itemize}[leftmargin=*]
  \item \code{-std=c++26} --- enable \cpp{}26. \maya{} requires it;
    \cpp{}20 or \cpp{}23 will not compile the DSL.
  \item \code{-Wall -Wextra} --- turn on most warnings. Treat
    warnings as bugs while learning.
  \item \code{-g} --- include debug information, so a stack trace
    in \code{gdb} or \code{lldb} points at your code.
  \item \code{-O2} --- ask for optimisation. Off by default; turn
    it on when you measure performance.
  \item \code{-I path/to/headers} --- add a directory to the
    include search path. \maya{} consumers pass
    \code{-I maya/include}.
  \item \code{-fsanitize=address,undefined} --- compile in run-time
    checks for memory bugs and undefined behaviour. Free safety
    net while learning.
\end{itemize}

For a one-liner sanity check that your compiler is recent enough:

\begin{lstlisting}[style=shell]
echo '#include <print>\nint main(){std::println("ok");}' \
  | g++-15 -std=c++26 -x c++ - -o ok && ./ok
\end{lstlisting}

If that prints \code{ok}, you can run every example in this book.

\section{Statements, blocks, control flow}
\label{sec:cpp:flow}

Before we go further into types, a quick survey of the syntactic
furniture every \cpp{} program is built from. If you have written
any imperative language --- Python, JavaScript, Java, C, Go ---
almost nothing here is new; if you have not, read carefully.

\subsection{Statements and blocks}

A \textbf{statement} is one unit of work. Most statements end with a
semicolon:

\begin{lstlisting}
int x = 0;
x = x + 1;
std::println("x is {}", x);
\end{lstlisting}

A \textbf{block} is a sequence of statements wrapped in braces. It
introduces a new scope --- names declared inside the block disappear
at the closing brace:

\begin{lstlisting}
int x = 0;
{
    int y = 5;      // y exists only inside this block
    x += y;
}
// y is gone here; x is 5
\end{lstlisting}

Function bodies, loop bodies, \code{if} branches --- all are blocks.

\subsection{\code{if} and \code{else}}

\begin{lstlisting}
if (x > 0) {
    std::println("positive");
} else if (x < 0) {
    std::println("negative");
} else {
    std::println("zero");
}
\end{lstlisting}

The condition must be a \code{bool} (or something that converts to
one --- an \code{int}, a pointer). \cpp{}17 added \textbf{if-init}:
you can declare a variable in the condition itself:

\begin{lstlisting}
if (auto it = map.find(key); it != map.end()) {
    use(it->second);
}
// it is not in scope here
\end{lstlisting}

The form \code{if (init; cond)} keeps the helper variable confined
to the \code{if}/\code{else} blocks. \maya{} uses this heavily.

\subsection{Loops: \code{while}, \code{do-while}, \code{for}}

\begin{lstlisting}
while (cond) {
    // body, repeated as long as cond is true
}

do {
    // body, runs at least once, then while-loops
} while (cond);

for (int i = 0; i < 10; ++i) {
    // classic C-style for: init; cond; step
}
\end{lstlisting}

The modern \textbf{range-for} (\cpp{}11) iterates over any container
or range:

\begin{lstlisting}
std::vector<int> xs = {1, 2, 3};
for (int x : xs)        std::println("{}", x);   // copy each element
for (int& x : xs)       ++x;                     // mutate in place
for (const auto& x : xs) std::println("{}", x);  // read-only, no copy
\end{lstlisting}

The \code{const auto\&} form is the most common in \maya{}: no copy,
no surprises.

\subsection{\code{break} and \code{continue}}

\code{break} exits the innermost loop. \code{continue} skips the
rest of the current iteration and goes to the next.

\begin{lstlisting}
for (int i = 0; i < 10; ++i) {
    if (i == 3) continue;   // skip 3
    if (i == 7) break;      // stop entirely
    std::println("{}", i);   // prints 0 1 2 4 5 6
}
\end{lstlisting}

\subsection{\code{switch}}

\begin{lstlisting}
switch (kind) {
    case Kind::Red:    handle_red();    break;
    case Kind::Green:  handle_green();  break;
    case Kind::Blue:   handle_blue();   break;
    default:           handle_other();  break;
}
\end{lstlisting}

The argument must be an integral or enum value. Each \code{case}
label falls through to the next unless terminated with \code{break}
(or \code{return}). \cpp{} accepts the fall-through; modern
compilers warn about unintended ones, and the attribute
\code{[[fallthrough]];} silences the warning for intentional cases.

In \maya{}, \code{switch} is used on small \code{enum class} kinds
(error kinds, key kinds). For closed sum types, \code{std::visit} on
a variant is preferred --- it forces exhaustiveness in a way
\code{switch} cannot.

\subsection{Expressions and statements}

Many things look like statements but are actually \textbf{expressions}
--- they produce a value. \code{x + 1}, \code{f(3)}, \code{a == b},
\code{x ? y : z} (the ternary conditional). Assignment is also an
expression; \code{(x = 5)} has value $5$. This rarely matters until
you meet a chained assignment or a condition like \code{while ((c =
read()))}.

\section{Built-in types and integer widths}
\label{sec:cpp:types}

C and \cpp{} have a long list of arithmetic types, mostly inherited
from 1970s hardware. You only need a few.

\begin{center}
\small
\begin{tabular}{lll}
\textbf{Type} & \textbf{Typical size} & \textbf{Use for} \\
\hline
\code{bool}            & 1 byte        & truth values \\
\code{char}            & 1 byte        & a byte / ASCII letter \\
\code{int}             & 4 bytes       & ordinary integers \\
\code{long}            & 4 or 8        & avoid; use a fixed-width type \\
\code{long long}       & 8             & avoid; use \code{std::int64\_t} \\
\code{float}           & 4             & single-precision floating-point \\
\code{double}          & 8             & default floating-point \\
\code{std::size\_t}    & pointer-sized & sizes, indices \\
\code{std::ptrdiff\_t} & pointer-sized & signed sizes, pointer differences \\
\code{std::uint8\_t}   & 1             & a byte you do arithmetic on \\
\code{std::int32\_t}   & 4             & an exactly-32-bit signed integer \\
\code{std::int64\_t}   & 8             & an exactly-64-bit signed integer \\
\end{tabular}
\end{center}

Two rules:

\begin{itemize}[leftmargin=*]
  \item Use \code{std::size\_t} for sizes and indices into
    containers. \code{std::vector::size()} returns it; comparing it
    against a signed \code{int} warns.
  \item Use the fixed-width types (\code{std::int32\_t}, etc.) when
    you care about the exact bit count --- file formats, hash
    functions, packed cell layouts.
\end{itemize}

\maya{}'s canvas cell, for example, is a \code{std::uint64\_t} ---
exactly 64 bits, the same on every platform. The diff pass loads it
into a SIMD register; that requires a known size.

\subsection{Signed vs unsigned}

\code{int} is signed; \code{unsigned int} is not. The difference
matters for two reasons:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Overflow.} Signed overflow is
    \emph{undefined behaviour}: the compiler may produce nonsense
    if you cause it. Unsigned overflow is defined: it wraps modulo
    $2^N$.
  \item \textbf{Comparison.} Mixing signed and unsigned in a
    comparison surprises everyone. \code{-1 < 0u} is
    \emph{false}: the \code{-1} is converted to \code{unsigned}
    first and becomes a huge positive number.
\end{enumerate}

Use signed types unless you genuinely need the wrap-around semantics
of unsigned --- bit twiddling, hashing, modular arithmetic, sizes.

\subsection{Character types}

\code{char} is one byte. Whether it is signed or unsigned is
implementation-defined --- never assume. \code{signed char} and
\code{unsigned char} explicitly specify. For UTF-8 byte buffers,
\code{std::uint8\_t} is the clearest choice; for ASCII text,
\code{char} is fine.

\cpp{} also has \code{char8\_t}, \code{char16\_t}, \code{char32\_t}
for wider Unicode --- you will see them rarely.

\subsection{Literals}

Number literals have suffixes that fix their type:

\begin{lstlisting}
42         // int
42u        // unsigned int
42L        // long
42LL       // long long
42ULL      // unsigned long long
3.14       // double
3.14f      // float
0x2A       // hex (= 42)
0b101010   // binary (= 42)
0'042      // digit separator allowed --- = 042 (octal = 34)
1'000'000  // a million, with separators for readability
\end{lstlisting}

String literals: \code{"hello"} is a \code{const char[6]} (the five
letters plus a trailing \code{\textbackslash 0}). \code{"hello"sv}
(with the \code{sv} suffix, after \code{using namespace
std::string\_view\_literals}) is a \code{std::string\_view}.

\section{Namespaces}
\label{sec:cpp:namespaces}

A \textbf{namespace} is a labelled container of names. It exists to
prevent collisions: every library can put its types inside its own
namespace, and no one tramples on anyone else's identifiers.

\begin{lstlisting}
namespace mylib {
    int add(int a, int b) { return a + b; }
    struct Point { double x, y; };
}

int main() {
    int n = mylib::add(2, 3);
    mylib::Point p{1.0, 2.0};
}
\end{lstlisting}

The scope-resolution operator \code{::} reaches into a namespace.
\code{std::vector} means ``\code{vector} inside the \code{std}
namespace''. The standard library lives entirely in \code{std}.

Namespaces nest:

\begin{lstlisting}
namespace maya {
    namespace ansi {
        constexpr std::string_view bold = "\\x1b[1m";
    }
}

// or, equivalently (C++17):
namespace maya::ansi {
    constexpr std::string_view bold = "\\x1b[1m";
}
\end{lstlisting}

You use it as \code{maya::ansi::bold}.

\subsection{\code{using} declarations}

Typing \code{maya::dsl::} in front of every DSL identifier would be
tedious. Three ways to abbreviate:

\begin{lstlisting}
using maya::dsl::Bold;          // bring just one name into scope
using namespace maya::dsl;      // bring every name into scope
namespace md = maya::dsl;       // alias the whole namespace
\end{lstlisting}

\code{using namespace X} is convenient at the top of a \code{.cpp}
file but \textbf{never} at namespace scope in a header --- you would
leak every name into every file that includes the header.

Maya's public namespaces:

\begin{itemize}[leftmargin=*]
  \item \code{maya} --- public API (\code{print}, \code{run},
    \code{Element}, \code{Signal}, \code{Cmd}).
  \item \code{maya::dsl} --- the compile-time DSL pieces
    (\code{t<"...">}, \code{Bold}, \code{Fg<...>}, \code{v},
    \code{h}, the pipe operators).
  \item \code{maya::ansi} --- raw escape sequence constants and
    builders.
  \item \code{maya::widget} --- the widget catalogue.
\end{itemize}

Example programs begin with \code{using namespace maya; using namespace
maya::dsl;} for brevity.

\section{Initialisation: the many forms}
\label{sec:cpp:init}

\cpp{} has \emph{four} syntaxes for initialising a variable. This is
the single most baffling piece of beginner \cpp{}. We will lay them
out, and then give you a rule.

\begin{lstlisting}
int a;            // default initialisation: uninitialised garbage!
int b = 5;        // copy initialisation
int c(5);         // direct initialisation
int d{5};         // direct list initialisation ("brace init")
auto e = 5;       // auto + copy form
auto f = int{5};  // auto + brace form
\end{lstlisting}

For a built-in \code{int}, the second through fifth are equivalent.
For classes, the difference matters in two places.

\subsection{Default initialisation can leave garbage}

\code{int a;} declares \code{a} but does \emph{not} give it a value.
Reading \code{a} before you assign to it is undefined behaviour.
This is one of the most common beginner bugs.

Use \code{int a\{\};} or \code{int a = 0;} to be explicit.
\code{T x\{\}} \textbf{value-initialises} \code{x} for any \code{T}
--- zero for built-ins, default-constructor for classes.

\subsection{Brace init avoids narrowing}

\begin{lstlisting}
int    a = 3.7;     // ok, silently truncates to 3
int    b{3.7};      // ERROR: narrowing conversion not allowed
\end{lstlisting}

Braces forbid lossy conversions. This catches real bugs.

\subsection{The ``most vexing parse''}

\begin{lstlisting}
Widget w();         // looks like a default construction, but...
                    //   ... it is actually a FUNCTION DECLARATION
                    //   for a function w taking nothing,
                    //   returning Widget.
Widget w;           // correct default construction
Widget w{};         // correct, explicit
\end{lstlisting}

Use braces or empty parentheses where you really mean to construct.

\subsection{The rule}

For the rest of this book, the rule is: \textbf{prefer braces}.

\begin{lstlisting}
int            n{0};
std::string    s{"hello"};
std::vector<int> v{1, 2, 3};
Point          p{3.0, 4.0};
auto           m = std::map<int, std::string>{{1, "a"}, {2, "b"}};
\end{lstlisting}

The one common exception: when calling a constructor whose arguments
should \emph{not} be interpreted as a list, you sometimes need
parentheses. \code{std::vector<int> v\{10, 0\}} is a vector of two
elements $[10, 0]$; \code{std::vector<int> v(10, 0)} is a vector of
ten zeros. Read the API docs when this comes up.

\section{Values, references, pointers, \code{const}}
\label{sec:cpp:memory}

\cpp{} forces you to be precise about \emph{where} a piece of data
lives and \emph{who can change it}. This precision is what makes
\cpp{} fast and what makes beginners stumble. We will walk it slowly.

\subsection{Values}

A \textbf{value} is just data: \code{int x = 42;} creates a variable
\code{x} of type \code{int} holding the value $42$. The variable
owns its storage. When \code{x} goes out of scope (the block ends),
the storage is gone.

\begin{lstlisting}
int x = 42;          // x owns 4 bytes holding 42
int y = x;           // y owns its own 4 bytes, copy of x
y = 100;             // x is still 42
\end{lstlisting}

This is the \textbf{value semantics} model: every assignment is a
copy, every variable is its own thing. The same applies to classes:

\begin{lstlisting}
std::string a = "hello";
std::string b = a;   // b is its own string, copy of a\'s characters
b += " world";
// a == "hello", b == "hello world"
\end{lstlisting}

For a \code{std::string} this is a real copy of every character.
That sounds wasteful and sometimes is; we will see \code{std::move}
shortly as the way to opt out of copies when you no longer need the
original.

\subsection{References}

A \textbf{reference} is an alias for an existing value. It is \emph{not}
a separate variable; it is another name for the same storage.

\begin{lstlisting}
int x = 42;
int& r = x;        // r is an alias for x
r = 100;
// x is now 100
\end{lstlisting}

References must be bound when created and cannot be re-seated.
\code{int\& r;} alone is not legal --- a reference always names
something.

References are the standard way to pass large values into a function
without copying:

\begin{lstlisting}
void print_long(const std::string& s) {
    std::println("{}", s);     // s is the caller\'s string
}                                  // no copy, no destruction
\end{lstlisting}

The \code{const} qualifier on a reference promises ``I will not
modify the thing you let me look at''. \code{const T\&} is the
idiomatic way to receive read-only inputs of any non-trivial type.

\subsection{Pointers}

A \textbf{pointer} is a value that holds an address --- the location
of another value in memory. The syntax \code{T*} means ``pointer to a
\code{T}''. You dereference with \code{*}.

\begin{lstlisting}
int x = 42;
int* p = &x;         // p holds the address of x
*p = 100;            // write through the pointer
// x is now 100
\end{lstlisting}

Unlike references, pointers \emph{can} be re-seated, can be null
(\code{nullptr}), and can point at nothing meaningful (a real source
of bugs). \maya{} avoids raw pointers in user-facing APIs.
Internally, it uses them for things that genuinely model ``maybe
points to something'': linked lists, lookup tables, parent links
in a tree.

When you do see a pointer in \maya{}, prefer the modern alternatives:
\code{std::unique\_ptr<T>} (single owner, automatically destroyed),
\code{std::shared\_ptr<T>} (reference-counted shared owner),
\code{std::optional<T>} (a value that may or may not be present, no
heap allocation), \code{std::span<T>} (a non-owning view of a
contiguous range), \code{std::string\_view} (the same for character
data).

\subsection{\code{const} on values and references}

A \code{const} variable cannot be reassigned:

\begin{lstlisting}
const int n = 5;
// n = 10;          // error: cannot assign to const
\end{lstlisting}

A \code{const} reference promises read-only access:

\begin{lstlisting}
void f(const std::string& s) {
    // s.push_back('!');  // error: s is const here
    std::println("{}", s);
}
\end{lstlisting}

A \code{const} method on a class promises not to modify the object:

\begin{lstlisting}
struct Counter {
    int n = 0;
    int get() const { return n; }   // const method, ok to call on const objects
    void inc()       { ++n; }
};

const Counter c;
c.get();           // ok
// c.inc();        // error: cannot call non-const method on const object
\end{lstlisting}

\code{const}-correctness is a deceptively powerful discipline: it
lets the compiler check at compile time that your read-only paths
really are read-only.

\begin{recap}
A value owns its storage. A reference is an alias; a pointer holds
an address. \code{const} marks read-only at every level: variable,
reference, method.
\end{recap}

\section{Functions, overloads, default arguments}
\label{sec:cpp:functions}

A function takes inputs, may return an output, and may have side
effects:

\begin{lstlisting}
int square(int n) { return n * n; }
void greet(std::string_view name) {
    std::println("hi, {}", name);
}
\end{lstlisting}

The return type goes first, then the name, then the parameters in
parentheses. \code{void} means ``returns nothing''.

\subsection{Parameter passing}

A parameter declared by value (\code{int n}) is a fresh copy. A
parameter declared by reference (\code{const T\&}) shares storage
with the caller's value. The rule of thumb:

\begin{itemize}[leftmargin=*]
  \item Trivial small types (\code{int}, \code{double},
    \code{std::string\_view}, \code{std::span<T>}) --- by value.
    Copying is cheap.
  \item Larger types you only read --- by \code{const} reference.
    No copy.
  \item Types you intend to modify in place --- by non-\code{const}
    reference. The caller sees the change.
  \item Types you want to take ownership of --- by value (and let
    the caller \code{std::move} into the call).
\end{itemize}

\subsection{Overloading}

Multiple functions can share a name if their parameter lists differ.
This is \textbf{overloading}:

\begin{lstlisting}
void log(int n);
void log(double d);
void log(std::string_view s);

log(42);          // calls log(int)
log(3.14);        // calls log(double)
log("hello");     // calls log(string_view)
\end{lstlisting}

The compiler picks the best match. If two are equally good, you get
a compile error and you spell out the choice.

\subsection{Default arguments}

A parameter can have a default:

\begin{lstlisting}
void greet(std::string_view name, std::string_view greeting = "hi") {
    std::println("{}, {}", greeting, name);
}

greet("world");            // "hi, world"
greet("world", "hello");   // "hello, world"
\end{lstlisting}

Defaults are part of the declaration; later overloads of the same
name cannot redeclare them.

\section{\code{enum class}: typed enumerations}
\label{sec:cpp:enum}

An \textbf{enumeration} is a small named set of integer constants.
The old C-style form was \code{enum Color \{ Red, Green, Blue \};},
which defined three constants in the surrounding scope --- \code{Red},
\code{Green}, \code{Blue} --- of an implicit integer type. The names
leaked, and the values implicitly converted to and from \code{int}.

\cpp{}11 introduced \code{enum class}, which fixes both problems:

\begin{lstlisting}
enum class Color { Red, Green, Blue };

Color c = Color::Red;          // scoped: must write Color::
int   n = c;                   // ERROR: no implicit conversion
int   n = static_cast<int>(c); // ok
\end{lstlisting}

The enumerators are scoped inside the \code{Color} type. They do not
implicitly convert to or from integers. Two unrelated
\code{enum class}es are unrelated types.

You can fix the underlying integer width:

\begin{lstlisting}
enum class Kind : std::uint8_t { A, B, C };  // each value fits in a byte
\end{lstlisting}

Maya uses \code{enum class} everywhere a value is one of a small
set:

\begin{lstlisting}
enum class HttpErrorKind { Tls, Timeout, Refused, Status, Cancel };
enum class BorderStyle   { None, Square, Round, Double, Thick };
enum class FlexDirection { Row, Column };
enum class KeyKind       { Char, Enter, Tab, Backspace, Escape,
                           Up, Down, Left, Right, ... };
\end{lstlisting}

When the value is one of more than a handful of cases, or when each
case carries data, prefer \code{std::variant} (Section
\ref{sec:cpp:variant}). \code{enum class} is for plain tagged states.

\section{Type aliases: \code{using}}
\label{sec:cpp:aliases}

Long type names are tedious to repeat. The old C way was
\code{typedef}:

\begin{lstlisting}
typedef std::vector<std::pair<std::string, int>> ScoreList;
\end{lstlisting}

\cpp{}11 introduced a cleaner form, \code{using}:

\begin{lstlisting}
using ScoreList = std::vector<std::pair<std::string, int>>;
\end{lstlisting}

It reads left-to-right (``\code{ScoreList} is\dots'') and supports
templates, which \code{typedef} does not:

\begin{lstlisting}
template <typename T>
using Vec2 = std::vector<std::vector<T>>;

Vec2<int> matrix;   // std::vector<std::vector<int>>
\end{lstlisting}

Maya defines aliases for every public type that has a long template
expansion (\code{using HttpResult = std::expected<Response,
HttpError>;} for example). Prefer \code{using} over \code{typedef}
everywhere; \code{typedef} is in the language only for C
compatibility.

\section{Attributes and \code{noexcept}}
\label{sec:cpp:attributes}

\cpp{}'s \textbf{attributes} are double-bracketed annotations the
compiler reads to enable warnings, optimisations, or sanity checks.
They do not change semantics; ignoring them is always safe.

\begin{itemize}[leftmargin=*]
  \item \code{[[nodiscard]]} on a function says ``do not silently
    throw away my return value''. The compiler warns if you call
    the function and ignore the result.
  \item \code{[[noreturn]]} says ``this function never returns''
    (it throws, calls \code{std::abort}, loops forever). Lets the
    compiler treat code after the call as unreachable.
  \item \code{[[deprecated("reason")]]} marks an API as old; uses
    of it warn.
  \item \code{[[maybe\_unused]]} on a variable or parameter
    silences the ``unused'' warning when the variable is
    intentionally unused.
  \item \code{[[likely]]} / \code{[[unlikely]]} on a branch hints
    to the compiler which way the branch usually goes.
\end{itemize}

In \maya{} you will see \code{[[nodiscard]]} on almost every
function returning a value. Ignoring the result of
\code{ansi::fg(c)} is a bug; the attribute makes the bug a warning.

\code{noexcept} is not an attribute --- it is part of the function
signature, before the body:

\begin{lstlisting}
int  add(int a, int b) noexcept { return a + b; }
void f() noexcept { /* must not throw */ }
\end{lstlisting}

It promises the function will not throw an exception. The compiler
can generate slightly faster code for \code{noexcept} functions
(no unwinding tables needed at the call site). \code{std::move},
\code{std::swap}, and many standard-library operations behave better
on types whose key operations are \code{noexcept}; in particular,
\code{std::vector} reallocations move (not copy) elements only when
the move constructor is \code{noexcept}.

Rule: mark a function \code{noexcept} when you are sure it will not
throw. Mark destructors \code{noexcept} essentially always (the
language already implicitly does this).

\section{Classes, constructors, destructors}
\label{sec:cpp:classes}

A \textbf{class} (introduced with \code{class} or \code{struct} ---
the only difference is the default access level) is a user-defined
type that bundles data and behaviour.

\begin{lstlisting}
struct Point {
    double x;
    double y;

    double length() const {
        return std::sqrt(x*x + y*y);
    }
};

Point p{3.0, 4.0};
std::println("{}", p.length());   // 5
\end{lstlisting}

\code{struct} members are public by default; \code{class} members
are private by default. Most \maya{} types are \code{struct}.

\subsection{Constructors}

A \textbf{constructor} is a special function that runs when an object
is created. It has the same name as the class, no return type, and
may take parameters.

\begin{lstlisting}
struct File {
    FILE* fp;

    explicit File(const char* path) {
        fp = std::fopen(path, "r");
    }
};

File log{"app.log"};   // calls the constructor with "app.log"
\end{lstlisting}

\code{explicit} prevents implicit conversion: without it, you could
write \code{File f = "app.log";} and \cpp{} would silently call the
constructor. Mark single-argument constructors \code{explicit} unless
you genuinely want the implicit form.

\subsection{Member initialiser lists}

The preferred way to set members from a constructor is the
\textbf{initialiser list}, which appears between a colon and the
body:

\begin{lstlisting}
struct File {
    FILE* fp;

    explicit File(const char* path) : fp(std::fopen(path, "r")) {}
};
\end{lstlisting}

Members are initialised in the order they are declared in the class,
\emph{not} in the order they appear in the initialiser list. Modern
compilers warn if you write them in the wrong order; pay attention
to the warning.

\subsection{Destructors}

A \textbf{destructor} is the constructor's counterpart: it runs when
the object is about to be destroyed. Its name is the class name
prefixed with \code{\~{}}.

\begin{lstlisting}
struct File {
    FILE* fp;

    explicit File(const char* path) : fp(std::fopen(path, "r")) {}
    ~File() { if (fp) std::fclose(fp); }
};

void load() {
    File log{"app.log"};   // fopen runs here
    // ... use log.fp ...
}                              // ~File runs here, fclose runs automatically
\end{lstlisting}

The destructor runs at the end of the scope where the object was
created, even if the function returns early, even if an exception is
thrown. This is the foundation of the next idea.

\section{RAII --- the central pattern}
\label{sec:cpp:raii}

\textbf{Resource Acquisition Is Initialization}, or \textbf{RAII}, is
the single most important idiom in \cpp{}. It is the principle that
every resource your program holds --- memory, file descriptor,
terminal mode, mutex lock, network connection --- should be tied to
the lifetime of an object. The constructor acquires; the destructor
releases. You cannot forget to release because the language runs the
destructor for you.

The \code{File} class above is RAII: constructing a \code{File} opens
the file; the destructor closes it. There is no \code{close()}
method to forget to call.

\subsection{A \code{TerminalGuard} for raw mode}

Let us write the RAII type that fixes the bug from
Chapter~\ref{ch:terminal}, where forgetting to restore cooked mode
left the user's shell broken.

\begin{lstlisting}
class TerminalGuard {
    termios cooked_;
public:
    TerminalGuard() {
        tcgetattr(STDIN_FILENO, &cooked_);
        termios raw = cooked_;
        raw.c_lflag &= ~(ICANON | ECHO | ISIG | IEXTEN);
        raw.c_iflag &= ~(IXON | ICRNL | INPCK | ISTRIP);
        raw.c_oflag &= ~(OPOST);
        raw.c_cc[VMIN]  = 0;
        raw.c_cc[VTIME] = 0;
        tcsetattr(STDIN_FILENO, TCSAFLUSH, &raw);
    }

    ~TerminalGuard() {
        tcsetattr(STDIN_FILENO, TCSAFLUSH, &cooked_);
    }

    TerminalGuard(const TerminalGuard&)            = delete;
    TerminalGuard& operator=(const TerminalGuard&) = delete;
};

int main() {
    TerminalGuard guard;   // raw mode on
    // ... run a TUI ...
}                          // raw mode off, automatically
\end{lstlisting}

The two \code{= delete} lines forbid copying. Copying would let two
\code{TerminalGuard} objects fight over the terminal state. We will
see \code{= delete} again shortly.

\maya{}'s actual terminal driver is more elaborate (it also handles
the alt-screen, mouse mode, KKP push, signal restoration on crash),
but the shape is exactly this: a class whose lifetime is the
lifetime of the terminal being in TUI mode.

\begin{idea}
RAII: every resource is owned by an object. Acquire in the
constructor, release in the destructor. Forgetting becomes a
compile-time impossibility, not a runtime risk.
\end{idea}

\subsection{The rule of zero, three, five}

If a class manages a resource, you usually need to define five
special member functions: destructor, copy constructor, copy
assignment, move constructor, move assignment. This is the
\textbf{rule of five}.

If you do not manage any resource yourself --- you only hold other
RAII types as members (\code{std::string}, \code{std::vector},
\code{std::unique\_ptr}) --- you should define \emph{none} of them.
The compiler generates correct versions automatically. This is the
\textbf{rule of zero}, and it is the rule you should follow first.

The \textbf{rule of three} (\cpp{}98) is the pre-move-semantics
version: define destructor, copy constructor, copy assignment
together.

Most \maya{} classes follow the rule of zero. The few that hold raw
resources (the terminal guard, signal handlers, file descriptor
wrappers) follow the rule of five.

\section{Inheritance and \code{virtual} (and why \maya{} avoids them)}
\label{sec:cpp:inheritance}

A class can \textbf{inherit} from another, picking up its members:

\begin{lstlisting}
struct Shape {
    virtual double area() const = 0;   // pure virtual: subclasses must override
    virtual ~Shape() = default;        // virtual destructor: required for safe deletion
};

struct Circle : Shape {
    double r;
    double area() const override { return 3.14159 * r * r; }
};

struct Square : Shape {
    double s;
    double area() const override { return s * s; }
};

Shape* p = new Circle{2.0};
std::println("{}", p->area());   // 12.56 --- runtime dispatch to Circle::area
delete p;
\end{lstlisting}

Key words:

\begin{itemize}[leftmargin=*]
  \item \code{struct Circle : Shape} --- \code{Circle} inherits
    publicly from \code{Shape}; \code{Circle*} converts to
    \code{Shape*} implicitly.
  \item \code{virtual} on a method --- the method can be overridden
    by subclasses and dispatched at runtime through a vtable.
  \item \code{= 0} --- pure virtual: this class is
    \emph{abstract} (cannot be instantiated) and subclasses must
    implement.
  \item \code{override} --- a tag on the subclass method telling
    the compiler ``I am overriding a base method''. If the signature
    drifts and you no longer match, you get an error. Always
    write it.
  \item \code{virtual \~{}Shape()} --- destructor must be virtual
    if you ever \code{delete} a derived object through a base
    pointer.
\end{itemize}

This is the classic object-oriented model: polymorphism through
inheritance, runtime dispatch through vtables.

\subsection{Why \maya{} barely uses it}

Inheritance has costs:

\begin{itemize}[leftmargin=*]
  \item Every \code{virtual} call is an indirect jump through a
    function pointer in the vtable. The branch predictor cannot
    always foresee where it goes, and a missed prediction is many
    cycles wasted. The diff loop runs hundreds of thousands of
    times per frame; an indirect call per cell would be ruinous.
  \item Polymorphism through pointers requires heap allocation in
    most uses (\code{new Circle{...}}, \code{std::unique\_ptr<Shape>}).
    Allocation is a real cost.
  \item The set of subclasses is open: code adding a new \code{Shape}
    can appear in any translation unit, and no compiler check forces
    every consumer of \code{Shape} to handle it. Closed sum types
    do not have this problem.
\end{itemize}

Where a polymorphism model is needed, \maya{} prefers
\code{std::variant} + \code{std::visit}:

\begin{lstlisting}
using Shape = std::variant<Circle, Square, Triangle>;

double area(const Shape& s) {
    return std::visit(overload{
        [](const Circle& c)   { return 3.14159 * c.r * c.r; },
        [](const Square& s)   { return s.s * s.s; },
        [](const Triangle& t) { return 0.5 * t.base * t.height; },
    }, s);
}
\end{lstlisting}

No vtable, no heap allocation, \emph{closed} (adding a new
alternative makes every visitor a compile error until updated). This
is the pattern for \code{Element}, \code{Event}, \code{Cmd<Msg>},
\code{Sub<Msg>}, and every other ``one of these things'' in maya.

Inheritance survives in \maya{} only for two narrow uses:

\begin{itemize}[leftmargin=*]
  \item Empty base classes used as compile-time tags (CRTP, mixin
    patterns). No runtime cost.
  \item Standard-library types we cannot avoid (\code{std::exception}
    is a base class).
\end{itemize}

\begin{idea}
Closed sum types over runtime polymorphism. The compiler can check
exhaustiveness; the CPU does not pay for a vtable.
\end{idea}

\section{Operator overloading}
\label{sec:cpp:operators}

\cpp{} lets you teach operators new meanings for your types. This is
why you can write \code{a + b} where \code{a}, \code{b} are
\code{std::string}s; \code{std::string} has an \code{operator+} that
produces concatenation. \maya{}'s DSL leans on this hard ---
\code{|} is taught a brand-new meaning.

An overload of a binary operator can be a method or a free function:

\begin{lstlisting}
struct Vec2 {
    double x, y;

    // as a method (left-hand side is *this):
    Vec2 operator+(const Vec2& other) const {
        return {x + other.x, y + other.y};
    }
};

// as a free function:
Vec2 operator-(const Vec2& a, const Vec2& b) {
    return {a.x - b.x, a.y - b.y};
}

Vec2 a{1, 2}, b{3, 4};
Vec2 c = a + b;          // calls Vec2::operator+
Vec2 d = a - b;          // calls free operator-
\end{lstlisting}

Free-function form is preferred when the left operand could be of a
type you do not own (e.g.\ \code{int + Vec2}).

\subsection{Operators you can overload}

Most of them: \code{+ - * / \% \& | \^{} \~{} ! < > <= >= == != \&\&
|| << >> += -= ... [] () -> ->* , ++ --} (prefix and postfix), and
conversion operators. You cannot invent new operators or change
precedence or arity.

Four are particularly common:

\begin{itemize}[leftmargin=*]
  \item \code{operator()} --- the call operator. Makes the type
    callable: \code{f(x, y)} where \code{f} is an object. Lambdas
    use it; \code{std::function} uses it.
  \item \code{operator[]} --- subscript. \code{v[i]} on containers.
  \item \code{operator T()} --- a \emph{conversion operator}. Lets
    your type implicitly convert to \code{T}. This is how
    \code{StyTag<...>} becomes \code{Style} in the DSL.
  \item \code{operator|} --- bitwise-or for built-in integers, but
    \maya{} overloads it to mean ``compose this style / modifier
    onto the thing on the left''.
\end{itemize}

Example of \code{operator|} for a tiny style system:

\begin{lstlisting}
struct Style { bool bold = false; bool italic = false; };
struct BoldTag {};
struct ItalicTag {};

constexpr Style operator|(Style s, BoldTag)   { s.bold = true;   return s; }
constexpr Style operator|(Style s, ItalicTag) { s.italic = true; return s; }

constexpr Style red = Style{} | BoldTag{} | ItalicTag{};
\end{lstlisting}

Left-to-right associativity of \code{|} makes the chain read
naturally: \code{Style\{\}} is the start; each tag layers on a flag.
\maya{}'s DSL is the same pattern, scaled up.

\subsection{Conversion operators}

A \code{operator T() const} lets your type act like a \code{T}:

\begin{lstlisting}
struct Fahrenheit {
    double value;
    operator double() const { return value; }
};

Fahrenheit f{72.0};
double d = f;            // implicit conversion: d == 72.0
std::println("{}", f);   // also works
\end{lstlisting}

Mark conversions \code{explicit} unless you really want them
implicit. Implicit conversions are silent and surprising; explicit
ones force the caller to write \code{static\_cast<double>(f)}.

The key exception: in \maya{}'s DSL, the \code{StyTag} types
implicitly convert to \code{Style}, and that implicit conversion is
the whole reason \code{node | Bold} works without writing
\code{static\_cast}.

\section{Smart pointers}
\label{sec:cpp:smartptr}

Raw pointers (\code{T*}) have ambiguous semantics: do they own the
object? Are they allowed to be null? The standard library provides
three \textbf{smart pointer} types that make the answer explicit.

\subsection{\code{std::unique\_ptr<T>}}

A \code{std::unique\_ptr<T>} is a single-owner pointer. It owns the
\code{T}; when it goes out of scope, it deletes the \code{T}. It
cannot be copied (otherwise two owners), only moved.

\begin{lstlisting}
#include <memory>
std::unique_ptr<Buffer> p = std::make_unique<Buffer>(1024);

p->write("hi", 2);          // -> dereferences
(*p).flush();               // * dereferences

std::unique_ptr<Buffer> q = std::move(p);   // q owns it now
// p is now empty (nullptr)
// at end of scope, q's destructor deletes the Buffer
\end{lstlisting}

Use this as your default. If a function returns a
\code{unique\_ptr<T>}, the caller becomes the owner. If a function
takes a \code{unique\_ptr<T>} by value, the caller has handed
ownership over.

\subsection{\code{std::shared\_ptr<T>}}

A \code{std::shared\_ptr<T>} is a reference-counted shared owner.
Multiple \code{shared\_ptr}s can point at the same object; the
object is deleted when the last one goes out of scope.

\begin{lstlisting}
std::shared_ptr<Cache> c = std::make_shared<Cache>();
std::shared_ptr<Cache> d = c;   // both own it
std::println("{}", c.use_count());   // 2
\end{lstlisting}

Use this only when you genuinely need shared ownership across
threads or owners. The reference count is atomic and a real cost;
for a single-owner relationship use \code{unique\_ptr}.

\subsection{\code{std::weak\_ptr<T>}}

A non-owning observer of a \code{shared\_ptr}. Used to break
reference cycles. You rarely write one in \maya{}.

\subsection{When to use raw \code{T*}}

Not for ownership. Only as a non-owning \emph{view} into something
you know outlives the pointer: a parent pointer in a tree, an
element of a stable array, a function parameter that may be null.
For pure ``read this thing'' parameters, prefer \code{const T\&};
for optional inputs, prefer \code{std::optional<T>}.

\maya{} uses \code{unique\_ptr} for heap-allocated children of
variants where appropriate (\code{Element} children are stored in a
\code{std::vector<Element>} rather than \code{vector<unique\_ptr>}
because \code{Element} is itself a variant with value semantics).
The few raw pointers are non-owning parent links.

\section{\code{auto} and type deduction}
\label{sec:cpp:auto}

Writing out the type of every variable is fine for \code{int} and
\code{double}; it is exhausting for things like
\code{std::vector<std::pair<std::string, Element>>::const\_iterator}.
\cpp{}11 introduced \code{auto}: ``compiler, figure out the type from
the initialiser''.

\begin{lstlisting}
auto x = 42;                  // int
auto pi = 3.14;               // double
auto name = std::string{"hi"};// std::string
auto v = std::vector<int>{1, 2, 3};
for (auto& item : v) ++item;  // item is int&
\end{lstlisting}

Three rules of thumb:

\begin{itemize}[leftmargin=*]
  \item Plain \code{auto} drops references and \code{const}. If you
    write \code{auto x = some\_const\_ref();} you get a fresh copy.
  \item \code{auto\&} keeps references and \code{const}. Use this in
    range-\code{for} loops to avoid copying each element.
  \item \code{const auto\&} is the most common form for read-only
    iteration: \code{for (const auto\& cell : row) \{ ... \}}.
\end{itemize}

In \maya{}, \code{auto} is the default everywhere a type would be
verbose. Spell types out only when the spelling is short or when it
clarifies intent.

\section{Move semantics}
\label{sec:cpp:move}

We saw that assignment copies. For a \code{std::vector<int>} with a
million elements, copying is a million-element allocation plus a
million-element copy. If you do not need the original after, the
copy is pure waste.

Move semantics is the \cpp{}11 feature that lets you say ``transfer
the contents, do not copy''. A move is usually a pointer swap: the
new vector inherits the buffer; the old vector is left empty.

\begin{lstlisting}
std::vector<int> a = {1, 2, 3};
std::vector<int> b = std::move(a);   // O(1) --- b owns the buffer
// a is in a "valid but unspecified" state; do not read it.
\end{lstlisting}

Key points:

\begin{itemize}[leftmargin=*]
  \item \code{std::move(x)} does not actually move anything; it is a
    \emph{cast} that says ``treat \code{x} as a thing whose contents
    may be stolen''. The move actually happens in the constructor
    or assignment operator that receives the rvalue.
  \item After moving from \code{x}, you must not assume \code{x}
    holds anything specific. You can still destroy it (safely) and
    you can still assign a fresh value to it.
  \item Move is an \emph{optimisation}, not a semantic change. A
    correctly written program works whether \code{std::move} is
    present or not; \code{std::move} only changes how fast it runs.
\end{itemize}

The \textbf{rvalue reference} type \code{T\&\&} is how a function
opts in to receiving a move:

\begin{lstlisting}
void take(std::string s);              // by value: caller can move into it
void take_ref(const std::string& s);   // by const ref: read-only
void take_rref(std::string&& s);       // rvalue ref: must be moved in

std::string name = "alice";
take(name);              // copies name into the parameter
take(std::move(name));   // moves name into the parameter
// after this, name is moved-from
\end{lstlisting}

In \maya{} you see \code{std::move} most often in the runtime:
\code{Model} flows through \code{update(Model, Msg)} without being
copied, because \code{update} takes its \code{Model} by value and
the runtime moves the current model into the call.

\begin{warning}
Do not \code{std::move} a local on the return statement (\code{return
std::move(x);}). The compiler already moves return values
automatically; writing \code{std::move} explicitly disables a
separate optimisation called RVO (Return Value Optimisation) that is
usually \emph{better} than move.
\end{warning}

\section{Exceptions, and why \maya{} prefers \code{expected}}
\label{sec:cpp:exceptions}

An \textbf{exception} is \cpp{}'s built-in mechanism for signalling
errors out of band: you \code{throw} a value, and the runtime unwinds
the stack (destroying every local along the way) until it finds a
matching \code{catch} handler.

\begin{lstlisting}
#include <stdexcept>

int sqrt_or_throw(int n) {
    if (n < 0) throw std::runtime_error("negative");
    return /* ... */;
}

try {
    int r = sqrt_or_throw(x);
    use(r);
} catch (const std::runtime_error& e) {
    std::println("failed: {}", e.what());
}
\end{lstlisting}

The machinery is real:

\begin{itemize}[leftmargin=*]
  \item \code{throw e} constructs and throws an exception value.
    Anything copyable or movable can be thrown; convention is to
    throw a class deriving from \code{std::exception}.
  \item Stack unwinding runs the destructor of every local between
    the \code{throw} and the \code{catch}. This is why RAII
    matters: your resources get released even on errors.
  \item \code{catch (const T\&)} matches a thrown \code{T} or any
    derived type. The last \code{catch} can be \code{catch (...)}
    to catch everything (rarely useful except at the top of
    \code{main}).
\end{itemize}

Exception handling is cheap on the success path (modern
implementations use zero-cost-when-not-thrown tables) and expensive
when an exception is actually thrown.

\subsection{Why \maya{} avoids them for ordinary errors}

Three reasons:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Invisibility.} Looking at the signature
    \code{int parse(std::string\_view s)}, you cannot tell whether
    it might throw. The caller has to know the convention or read
    the body. \code{std::expected<int, ParseError>} writes the
    failure mode into the signature.
  \item \textbf{Closed errors.} Catching by base class means the
    caller cannot meaningfully switch on the failure mode without
    \code{dynamic\_cast}-ing. \code{std::expected} with an
    \code{enum class Kind} error type gives a finite, switchable
    list.
  \item \textbf{Cost on the wrong path.} Throwing is slow. A TUI
    that throws on every key press it doesn't recognise will feel
    sluggish.
\end{enumerate}

\maya{} still uses exceptions for truly exceptional things ---
out-of-memory, bugs in the runtime, panic conditions at top level.
For ordinary, expected failures (parse failed, file missing,
network timed out), it returns \code{std::expected}.

\begin{idea}
Throw on bugs, return on errors. Exceptions for ``the world is
broken''; \code{std::expected} for ``this might not work, and that
is fine''.
\end{idea}

\section{Templates}
\label{sec:cpp:templates}

A \textbf{template} is a recipe the compiler runs at compile time.
You write the recipe once; the compiler bakes a fresh function or
class for each set of arguments you give it.

\subsection{Function templates}

\begin{lstlisting}
template <typename T>
T add(T a, T b) { return a + b; }

int    x = add(1, 2);          // T = int
double y = add(1.5, 2.5);      // T = double
std::string s = add<std::string>("foo", "bar");  // T = std::string
\end{lstlisting}

\code{typename T} introduces a type parameter named \code{T}. The
function body uses \code{T} as if it were a type. The compiler
\emph{instantiates} the template once per distinct \code{T}: there
are separate \code{add<int>}, \code{add<double>}, and
\code{add<std::string>} functions in the final binary.

The template only compiles correctly if \code{T} actually supports
\code{operator+}. Try \code{add(File, File)} and you will get a long
error. Concepts (Section~\ref{sec:cpp:concepts}) fix this.

\subsection{Class templates}

\begin{lstlisting}
template <typename T>
struct Box {
    T value;
    T& get() { return value; }
};

Box<int>    a{42};
Box<double> b{3.14};
Box<std::string> c{"hi"};
\end{lstlisting}

Each instantiation is a different type. \code{Box<int>} and
\code{Box<double>} share no relationship --- you cannot assign one
to the other.

\code{std::vector<T>}, \code{std::optional<T>},
\code{std::unique\_ptr<T>}, \code{std::variant<T, U, ...>} are all
class templates.

\subsection{Multiple template parameters}

\begin{lstlisting}
template <typename K, typename V>
struct Pair {
    K first;
    V second;
};

Pair<std::string, int> ages[] = {{"alice", 30}, {"bob", 25}};
\end{lstlisting}

\subsection{Non-type template parameters (NTTPs)}

The lesser-known but crucial flavour. A template parameter does not
have to be a type --- it can be a \emph{value}, computed at compile
time:

\begin{lstlisting}
template <int N>
struct Array {
    int data[N];
    int size() const { return N; }
};

Array<8>  a;   // data is int[8]
Array<16> b;   // data is int[16]
// a and b are different types
\end{lstlisting}

\code{std::array<T, N>} works this way. The size is part of the
type, so \code{std::array<int, 4>} and \code{std::array<int, 8>} are
distinct types and the size is known at compile time.

The types of NTTPs were limited for a long time --- integers,
pointers, enums. Since \cpp{}20, NTTPs can be \textbf{structural
class types}: any literal type with all-public members and no
user-defined comparison oddities. This is the trick that powers
\maya{}'s DSL.

\subsection{NTTPs of structural type --- the climax}

We want to write \code{t<"Hello">} and have the string become part of
the type. The straight path is to define a small ``string of fixed
length'' type and use it as an NTTP:

\begin{lstlisting}
template <std::size_t N>
struct Str {
    char data[N]{};

    consteval Str(const char (&s)[N]) {
        for (std::size_t i = 0; i < N; ++i) data[i] = s[i];
    }
};
\end{lstlisting}

Line by line:

\begin{itemize}[leftmargin=*]
  \item \code{template <std::size\_t N>}: a class template
    parameterised by an integer.
  \item \code{char data[N]\{\}}: a fixed-size array of \code{N}
    characters, zero-initialised.
  \item \code{consteval Str(const char (\&s)[N])}: a constructor that
    \emph{must} run at compile time (\code{consteval}) and takes a
    \emph{reference to a literal array of \code{N} chars}. The
    array-reference syntax \code{(\&s)[N]} captures the length of
    the literal --- the compiler picks \code{N} for you.
  \item The loop copies the string into \code{data}.
\end{itemize}

Now we can use \code{Str} as an NTTP:

\begin{lstlisting}
template <Str S>
struct Text {
    static constexpr std::string_view text() {
        return {S.data};
    }
};

Text<"Hello"> a;     // S = Str{"Hello\\0"}, baked into the type
Text<"World"> b;     // distinct type
\end{lstlisting}

\code{Text<"Hello">} and \code{Text<"World">} are different types.
The \emph{bytes} of the string are part of the type identity. This
is exactly how \maya{}'s \code{t<"...">} works, and it is the
foundation of the whole compile-time DSL we tour in
Chapter~\ref{ch:dsl}.

\begin{idea}
NTTPs let you encode \emph{values} into the type system. \maya{}
encodes text, colours, paddings, borders, and style flags this way.
Once a value is in a type, the compiler can check rules between
values at compile time.
\end{idea}

\subsection{Templates and the \code{<>} syntax}

When you see \code{Foo<X, Y>} in a \cpp{} expression, the angle
brackets are template-argument syntax, not less-than / greater-than.
The ambiguity is one of the language's papercuts; the compiler
usually figures it out from context, and where it cannot, you write
\code{template} explicitly: \code{x.template foo<int>()}.

You will see \code{t<"Hi">}, \code{Fg<255, 100, 100>}, \code{pad<1>},
\code{border\_<Round>} throughout \maya{}. They are all template
instantiations.

\section{Parameter packs and fold expressions}
\label{sec:cpp:packs}

A \textbf{variadic template} accepts any number of template arguments.
The declaration looks like a pack of dots:

\begin{lstlisting}
template <typename... Ts>
struct Tuple { /* ... */ };

Tuple<int, std::string, double> t;
\end{lstlisting}

\code{typename... Ts} introduces a \textbf{parameter pack} named
\code{Ts}. Inside the template, \code{Ts...} expands the pack ---
likewise for function-parameter packs:

\begin{lstlisting}
template <typename... Args>
void print_all(const Args&... args) {
    // args is a function-parameter pack
    // expanding (args)... yields a comma-separated list of arguments
}

print_all(1, "hi", 3.14);
\end{lstlisting}

The expansion syntax \code{pattern...} repeats the \code{pattern}
once per element of the pack, comma-separated.

\subsection{Fold expressions}

Writing recursion-style overloads to walk a pack used to be the only
way to consume it. \cpp{}17 added \textbf{fold expressions}: ask the
compiler to fold the pack with a binary operator.

\begin{lstlisting}
template <typename... Args>
auto sum(Args... args) {
    return (args + ...);     // unary right fold over +
}

sum(1, 2, 3, 4);              // 10 = ((1 + (2 + (3 + 4))))
\end{lstlisting}

Four forms exist (\code{(... op pack)},
\code{(pack op ...)}, \code{(init op ... op pack)},
\code{(pack op ... op init)}), but you usually just need one of:

\begin{lstlisting}
(args + ...)               // sum
(args, ...)                // evaluate left-to-right, throw away
(os << ... << args)        // chained <<: equivalent to os << a << b << c
((std::println("{}", args)), ...);  // print each, in order
\end{lstlisting}

The last one --- a fold over comma --- is the idiomatic ``do this
once per pack element''.

\subsection{The \code{overload} helper, revisited}

With packs in hand, the \code{overload} struct from earlier makes
full sense:

\begin{lstlisting}
template <typename... Fs>
struct overload : Fs... {
    using Fs::operator()...;
};
template <typename... Fs> overload(Fs...) -> overload<Fs...>;
\end{lstlisting}

\begin{itemize}[leftmargin=*]
  \item \code{template <typename... Fs>}: \code{Fs} is a pack of
    types --- the lambda types you pass in.
  \item \code{struct overload : Fs...}: inherit from every type in
    the pack. The pack expands into a comma-separated base list.
  \item \code{using Fs::operator()...}: bring each base's
    \code{operator()} into the combined class. This is a
    \textbf{using-declaration pack expansion}.
  \item The deduction guide \code{overload(Fs...) -> overload<Fs...>}
    lets you write \code{overload\{lambda1, lambda2\}} without
    spelling out the lambda types (which are unnameable anyway).
\end{itemize}

The result: \code{overload\{f, g, h\}} is a single object whose
\code{operator()} resolves to whichever of \code{f}, \code{g}, or
\code{h} matches the call. Combined with \code{std::visit}, it gives
you pattern matching on variants.

\section{\code{constexpr}, \code{consteval}, \code{constinit}}
\label{sec:cpp:constexpr}

A \textbf{constant expression} is one the compiler can evaluate at
translation time, before the program runs. \cpp{} has three
keywords governing this:

\begin{itemize}[leftmargin=*]
  \item \code{constexpr} on a function or variable: \emph{may} be
    constant-evaluated; otherwise behaves like a normal function or
    variable.
  \item \code{consteval} on a function: \emph{must} be
    constant-evaluated. Calling it at runtime is a compile error.
  \item \code{constinit} on a variable: must be initialised by a
    constant expression, but the variable itself is otherwise
    mutable. Used to forbid the ``static initialisation order
    fiasco''.
\end{itemize}

Example of \code{constexpr}:

\begin{lstlisting}
constexpr int square(int n) { return n * n; }

constexpr int sixteen = square(4);   // evaluated at compile time
int runtime_n = 7;
int forty_nine = square(runtime_n);  // also legal, evaluated at runtime
\end{lstlisting}

Example of \code{consteval}:

\begin{lstlisting}
consteval int square_only_ct(int n) { return n * n; }

constexpr int a = square_only_ct(4);  // ok
int runtime_n = 7;
// int b = square_only_ct(runtime_n); // error: must be constant-evaluated
\end{lstlisting}

\maya{}'s DSL is heavy on \code{constexpr} and \code{consteval}. The
entire chain \code{t<"Hi"> | Bold | Fg<255, 80, 80>} produces a
\code{constexpr} value --- every \code{operator|} along the way is
\code{constexpr}. Conversion to the runtime \code{Element} (which
may allocate) happens only when you call \code{.build()}.

\section{Concepts and \code{requires}}
\label{sec:cpp:concepts}

A template constrains its arguments implicitly: writing \code{a + b}
in a template body means \code{T} must support \code{operator+}, but
the constraint is invisible until you try to instantiate the
template with a wrong type and get a wall of compiler errors.

\cpp{}20 \textbf{concepts} make those constraints explicit and
named:

\begin{lstlisting}
template <typename T>
concept Addable = requires(T a, T b) {
    { a + b } -> std::convertible_to<T>;
};

template <Addable T>
T add(T a, T b) { return a + b; }

add(1, 2);                // ok, int is Addable
// add(File{...}, File{...});  // clean error: File does not satisfy Addable
\end{lstlisting}

The \code{requires} expression lists statements that must be valid
for \code{T}. \code{\{ a + b \} -> std::convertible\_to<T>} means
``\code{a + b} must compile, and its result must convert to
\code{T}''. If a caller passes a \code{T} that does not satisfy
\code{Addable}, the compiler points at the constraint, not at
something seventeen layers deep inside the template body.

\subsection{Maya's concepts}

Maya defines several concepts in \file{core/concepts.hpp}. Three of
the important ones:

\begin{lstlisting}
// (paraphrased from maya source)

template <typename T>
concept Node = requires(const T& t) {
    { t.build() } -> std::convertible_to<Element>;
};

template <typename P>
concept Program = requires(typename P::Model m, typename P::Msg msg) {
    typename P::Model;
    typename P::Msg;
    { P::init() };
    { P::update(m, msg) };
    { P::view(m) }       -> std::convertible_to<Element>;
    { P::subscribe(m) };
};
\end{lstlisting}

\code{Node} says ``has a \code{.build()} that returns something
convertible to \code{Element}''. Every DSL piece satisfies it.
\code{Program} says ``has the four Elm functions with the right
signatures''. \code{run<P>()} requires \code{Program<P>}; if your
program is missing a function, the compiler tells you which one.

\begin{idea}
Concepts are documentation that the compiler checks. They turn
``this template requires the type to do X'' from a comment into a
compile-time predicate.
\end{idea}

\section{\code{std::variant} and \code{std::visit}}
\label{sec:cpp:variant}

A \code{std::variant<A, B, C>} is a value that holds \emph{exactly
one} of \code{A}, \code{B}, or \code{C}. It is the \cpp{} closed sum
type --- the same construct functional languages call an algebraic
data type.

\begin{lstlisting}
#include <variant>
std::variant<int, std::string, double> v;
v = 42;          // holds an int
v = "hello";     // now holds a string
v = 3.14;        // now holds a double
\end{lstlisting}

You inspect a variant with \code{std::visit}, which dispatches to
the overload matching the current alternative. The idiomatic helper
is \code{overload}, a small struct that combines multiple lambdas
into a single callable:

\begin{lstlisting}
template <typename... Fs> struct overload : Fs... { using Fs::operator()...; };
template <typename... Fs> overload(Fs...) -> overload<Fs...>;

std::visit(overload{
    [](int n)               { std::println("int  : {}", n); },
    [](const std::string& s){ std::println("str  : {}", s); },
    [](double d)            { std::println("real : {}", d); },
}, v);
\end{lstlisting}

The \code{overload} template uses two features:

\begin{itemize}[leftmargin=*]
  \item \code{Fs...} is a \textbf{parameter pack} --- a variadic
    template, accepting any number of types.
  \item Inheriting from each one and using \code{using
    Fs::operator()...} brings every lambda's \code{operator()} into
    the combined class. \code{overload\{f, g, h\}} is then a single
    object with three call-operator overloads.
\end{itemize}

This is enough \cpp{} sorcery to write off as a one-liner you copy.
It is in \file{maya/core/overload.hpp}.

\subsection{Maya's variants}

Maya uses \code{std::variant} relentlessly:

\begin{itemize}[leftmargin=*]
  \item \code{Element} is
    \code{std::variant<BoxElement, TextElement, ElementList>}.
  \item \code{Event} is
    \code{std::variant<Key, Mouse, Resize, Paste, ...>}.
  \item \code{Cmd<Msg>} is a variant over every effect the runtime
    knows: \code{None}, \code{Quit}, \code{Batch}, \code{After},
    \code{SetTitle}, \code{WriteClipboard}, \code{Task}, ...
  \item \code{Sub<Msg>} is a variant over event sources.
\end{itemize}

Whenever you see a \code{std::visit} in \maya{}, it is the dispatch
hub for one of these types.

\begin{warning}
If you add a new alternative to a variant and forget to handle it in
a visitor, the visitor stops compiling. This is a feature: it makes
refactors safe. Embrace it; do not paper over it with a generic
fallback.
\end{warning}

\section{Modern syntax sugar you will see everywhere}
\label{sec:cpp:sugar}

A handful of small \cpp{}17 / 20 / 23 features change what idiomatic
code \emph{looks} like. They are not deep, but a beginner who has
never seen them will stumble. We collect them here.

\subsection{Structured bindings}

Given a struct, tuple, or pair, you can destructure it into named
variables in one line:

\begin{lstlisting}
std::pair<int, std::string> p{42, "hi"};
auto [n, s] = p;          // n = 42, s = "hi"

std::map<int, std::string> m;
for (const auto& [key, value] : m) {
    std::println("{} -> {}", key, value);
}

struct Point { double x, y; };
Point pt{3.0, 4.0};
auto [x, y] = pt;
\end{lstlisting}

The names are introduced as if by separate declarations.
\code{auto\&} and \code{const auto\&} work too: \code{auto\& [k, v]
: m} gives references, no copies. Maya uses structured bindings in
the runtime everywhere a function returns a \code{std::pair} ---
including \code{P::init()} and \code{P::update(m, msg)}.

\subsection{Designated initialisers}

For aggregates (structs with all-public fields), you can name fields
in the initialiser:

\begin{lstlisting}
struct Config {
    int  width  = 80;
    int  height = 24;
    bool mouse  = true;
};

Config a{};                              // all defaults
Config b{.width = 100};                  // override one
Config c{.width = 100, .height = 30};    // override two
\end{lstlisting}

The fields must appear in declaration order; you can skip fields
(they get their defaults). This produces self-documenting call sites:

\begin{lstlisting}
run<MyProgram>({.title = "counter", .alt_screen = true, .mouse = false});
\end{lstlisting}

Maya's \code{RunConfig} is exactly this shape; every example program
passes a designated-initialiser literal.

\subsection{Class template argument deduction (CTAD)}

You normally write \code{std::vector<int> v\{1, 2, 3\}}. \cpp{}17
lets you drop the \code{<int>} when the arguments make it obvious:

\begin{lstlisting}
std::vector v{1, 2, 3};                  // deduces vector<int>
std::pair  p{42, std::string{"hi"}};     // deduces pair<int, string>
std::array a{1, 2, 3, 4};                // deduces array<int, 4>
\end{lstlisting}

When you write your own class template, you can give it a
\textbf{deduction guide} (we saw one for \code{overload}). For most
user-defined classes the compiler deduces fine without one.

\subsection{\code{auto} return types}

A function's return type can be deduced from its return statements:

\begin{lstlisting}
auto add(int a, int b) {
    return a + b;       // return type deduced to int
}

auto first(const std::vector<int>& v) -> int {  // trailing return type
    return v.front();
}
\end{lstlisting}

The trailing-return-type syntax (\code{auto f(...) -> T}) is useful
when the return type depends on the parameters in a way the leading
position cannot reach.

For templates and lambdas, \code{auto} return is the default; the
compiler infers from the body. For \maya{}'s public API, return
types are usually written out for clarity.

\subsection{\code{constexpr if}}

Inside a template, \code{if constexpr (cond)} is a branch evaluated
at compile time: the false arm is not even instantiated.

\begin{lstlisting}
template <typename T>
void print(const T& x) {
    if constexpr (std::is_integral_v<T>) {
        std::println("int: {}", x);
    } else if constexpr (std::is_floating_point_v<T>) {
        std::println("float: {}", x);
    } else {
        std::println("other: {}", x);
    }
}
\end{lstlisting}

With plain \code{if}, both arms would need to compile for every
\code{T} (which would fail when \code{T} is not, say,
formattable). With \code{if constexpr}, only the live arm is
instantiated.

\maya{}'s DSL uses \code{if constexpr} extensively to specialise
behaviour by node kind without writing separate overloads.

\subsection{\code{nullptr}, not \code{NULL}}

\cpp{} 0 (= \code{NULL}) is an integer literal in disguise; it does
not type-check well. \cpp{}11 added \code{nullptr}, a value of type
\code{std::nullptr\_t} that converts to any pointer type but not to
\code{int}. Always write \code{nullptr}.

\section{\code{std::expected<T, E>}}
\label{sec:cpp:expected}

\cpp{}23's \code{std::expected<T, E>} is a sum type for fallible
operations: either a success of type \code{T} or an error of type
\code{E}. It is \maya{}'s preferred alternative to throwing
exceptions for ``failure is expected, just rare'' cases.

\begin{lstlisting}
#include <expected>
#include <charconv>

std::expected<int, std::string> parse_int(std::string_view s) {
    int n;
    auto [end, ec] = std::from_chars(s.data(), s.data() + s.size(), n);
    if (ec != std::errc{}) return std::unexpected("not an int");
    return n;
}

if (auto r = parse_int("42")) {
    std::println("value: {}", *r);
} else {
    std::println("error: {}", r.error());
}
\end{lstlisting}

The API at a glance:

\begin{itemize}[leftmargin=*]
  \item \code{(bool)r} --- true if success.
  \item \code{*r} or \code{r.value()} --- the success value.
  \item \code{r.error()} --- the error value.
  \item \code{std::unexpected(e)} --- factory for the error case.
\end{itemize}

In \maya{} the error type is usually a small struct:

\begin{lstlisting}
enum class HttpErrorKind { Tls, Timeout, Refused, Status, Cancel };
struct HttpError {
    HttpErrorKind kind;
    int           http_status;
    std::string   detail;
};
using HttpResult = std::expected<Response, HttpError>;
\end{lstlisting}

The caller switches on \code{kind} without parsing strings --- a
proper closed sum of failure modes. This is what we call a
``refined'' error type.

\section{Lambdas and capture}
\label{sec:cpp:lambdas}

A \textbf{lambda} is an anonymous function object. Under the hood,
the compiler synthesises a nameless class with an \code{operator()}
that runs the lambda body.

\begin{lstlisting}
auto add = [](int a, int b) { return a + b; };
int n = add(3, 4);    // 7
\end{lstlisting}

The square brackets at the front are the \textbf{capture list}: they
list names from the surrounding scope the lambda should remember.

\begin{lstlisting}
int x = 10;

auto by_value = [x](int n) { return x + n; };   // copy of x
auto by_ref   = [&x](int n){ return x + n; };   // reference to x

x = 99;
by_value(0);   // 10  --- the lambda has the old copy
by_ref(0);     // 99  --- the lambda sees the new x
\end{lstlisting}

Shorthand captures:

\begin{lstlisting}
[=]        // capture every used name by value
[&]        // capture every used name by reference
[=, &out]  // mostly by value, but out by reference
[&, n]     // mostly by reference, but n by value
\end{lstlisting}

Default captures (\code{[=]} and \code{[\&]}) are convenient but
dangerous in long-lived lambdas; you can easily capture a reference
to a local that has gone out of scope. Prefer explicit capture lists.

\subsection{Init-captures and moving into a lambda}

\cpp{}14 added \textbf{init-captures}: \code{[name = expression]}
introduces a new name local to the lambda with any initialiser. This
is how you capture a move-only type into a lambda:

\begin{lstlisting}
auto buf = std::make_unique<Buffer>();
auto cb  = [buf = std::move(buf)] {
    buf->flush();
};
// outer buf is moved-from; only the lambda owns the buffer.
\end{lstlisting}

\subsection{Storing lambdas: \code{std::function} vs.\ \code{std::move\_only\_function}}

A lambda's type is unique to it; you cannot write the type by hand.
To store a lambda in a class member or container, you need a
type-erased wrapper:

\begin{itemize}[leftmargin=*]
  \item \code{std::function<R(Args...)>} (\cpp{}11) holds any
    callable matching the signature, but the callable must be
    \emph{copyable}. That excludes lambdas capturing move-only state
    (\code{std::unique\_ptr}, sockets, file handles).
  \item \code{std::move\_only\_function<R(Args...)>} (\cpp{}23)
    requires only \emph{movable}. \maya{}'s signal callbacks and
    \code{Cmd<Msg>::Task} use this; effect descriptions routinely
    carry move-only state.
\end{itemize}

\section{Standard-library bits you will see}
\label{sec:cpp:stdlib}

The standard library is large. A non-exhaustive list of names that
appear in \maya{} and what they are:

\begin{itemize}[leftmargin=*]
  \item \code{std::string\_view} --- a non-owning view of a
    contiguous \code{char} sequence. Two pointers under the hood.
    Cheap to pass; does not own the bytes; the underlying data must
    outlive the view.
  \item \code{std::span<T>} --- the same idea for any contiguous
    range of \code{T}. \code{std::span<const Cell>} is how the
    renderer receives a row.
  \item \code{std::array<T, N>} --- a fixed-size array; size is a
    template argument. Same memory layout as \code{T[N]} but with
    proper value semantics and methods.
  \item \code{std::vector<T>} --- the dynamic-array workhorse.
    Heap-allocated buffer that grows.
  \item \code{std::optional<T>} --- either a \code{T} or nothing. Use
    when ``absent'' is a real state but you do not need a richer
    error.
  \item \code{std::format} / \code{std::print} / \code{std::println}
    --- \cpp{}20/23 formatted output, like Python's f-strings.
  \item \code{std::source\_location} --- a value carrying the file,
    line, and function name where it was constructed. \maya{} uses
    it for stable cache identifiers.
  \item \code{std::chrono} --- duration and time-point types.
    \code{std::chrono::milliseconds(50)} reads cleanly.
  \item \code{std::atomic<T>} --- lock-free operations on a value
    shared across threads.
  \item \code{std::jthread} --- a thread that joins on destruction
    (no leaked threads on early return). \cpp{}20.
  \item \code{std::ranges} / \code{std::views} --- lazy, composable
    algorithms over sequences (\code{xs | views::transform(...) |
    views::filter(...)}).
\end{itemize}

Learn the views as you meet them; the ranges library is large and you
do not need it all at once.

\section{Reading real \maya{} code}
\label{sec:cpp:reading}

With the toolkit in hand, let us read a fragment of actual \maya{}.
This is from the style header; it declares the tag types behind
\code{Bold}, \code{Fg<R, G, B>}, and friends.

\begin{lstlisting}
template <CTStyle V>
struct StyTag {
    constexpr operator Style() const noexcept { return V.runtime(); }
};

inline constexpr StyTag<CTStyle::bold()>   Bold{};
inline constexpr StyTag<CTStyle::italic()> Italic{};

template <std::uint8_t R, std::uint8_t G, std::uint8_t B>
inline constexpr StyTag<CTStyle::fg(R, G, B)> Fg{};
\end{lstlisting}

Line by line:

\begin{itemize}[leftmargin=*]
  \item \code{template <CTStyle V>}: \code{StyTag} is a class
    template parameterised by an NTTP \code{V} of type
    \code{CTStyle}. (\code{CTStyle} is a structural class type
    --- another struct defined elsewhere --- carrying a compile-time
    description of a style: which attributes are on, the colours,
    and so on.)
  \item \code{struct StyTag \{ ... \}}: the type has no data
    members. It is a \emph{tag} type --- empty at runtime, distinct
    at compile time. Two \code{StyTag<X>}'s with different \code{X}
    are different types.
  \item \code{constexpr operator Style() const noexcept}: an
    implicit conversion from \code{StyTag<V>} to the runtime
    \code{Style} type. \code{constexpr}, so it can run at compile
    time; \code{noexcept} promises it does not throw. \code{const}
    on the trailing position says the conversion does not modify
    the object.
  \item \code{return V.runtime();}: ask the compile-time style for
    its runtime form. \code{V} is the NTTP, available inside the
    class.
  \item \code{inline constexpr StyTag<CTStyle::bold()> Bold\{\};}: a
    \code{constexpr} variable of type
    \code{StyTag<CTStyle::bold()>}, initialised with default braces.
    \code{inline} on a variable in a header lets it appear in
    multiple translation units without violating ODR. \code{Bold} is
    the value you write in user code.
  \item \code{Fg<R, G, B>}: a \emph{variable template} ---  a
    template that yields a variable, not a function or class. The
    expression \code{Fg<255, 80, 80>} instantiates the template with
    those three integers, producing a unique \code{StyTag<...>}
    value.
\end{itemize}

If you can read that fragment now and feel the pieces in place ---
NTTP, class template, \code{constexpr}, \code{operator T()},
variable template, the implicit conversion that lets \code{Bold}
appear in places that want a \code{Style} --- you can read the rest
of \maya{}'s headers. From here on, the rest of the book is mostly
about what each file \emph{does}, not how its syntax decodes.

\subsection{A second worked fragment: \code{Cmd<Msg>}}

For practice, here is a paraphrased shape of \maya{}'s
\code{Cmd<Msg>} from \file{core/cmd.hpp}. \code{Cmd<Msg>} is the
type of effect descriptions the runtime knows how to perform.

\begin{lstlisting}
template <typename Msg>
class Cmd {
public:
    struct None {};
    struct Quit {};
    struct After {
        std::chrono::milliseconds delay;
        Msg                       msg;
    };
    struct SetTitle      { std::string title; };
    struct WriteClipboard{ std::string content; };
    struct Batch         { std::vector<Cmd> cmds; };
    struct Task {
        std::move_only_function<void(std::function<void(Msg)>)> run;
    };

    using Variant = std::variant<None, Quit, After, SetTitle,
                                 WriteClipboard, Batch, Task /*, ...*/>;

    Cmd() : v_(None{}) {}
    Cmd(Variant v) : v_(std::move(v)) {}

    [[nodiscard]] static Cmd none() { return Cmd{}; }
    [[nodiscard]] static Cmd quit() { return Cmd{Quit{}}; }
    [[nodiscard]] static Cmd set_title(std::string s) {
        return Cmd{SetTitle{std::move(s)}};
    }
    // ... and so on for every effect

    template <typename Visitor>
    auto visit(Visitor&& vis) const {
        return std::visit(std::forward<Visitor>(vis), v_);
    }

private:
    Variant v_;
};
\end{lstlisting}

Line by line:

\begin{itemize}[leftmargin=*]
  \item \code{template <typename Msg>}: \code{Cmd} is a class
    template parameterised by the program's message type, so each
    program has its own \code{Cmd} type carrying its own
    \code{Msg}s.
  \item \code{struct None \{\}; struct Quit \{\}; ...}: each
    effect is a tiny inner struct. Empty structs (\code{None},
    \code{Quit}) have no data; others carry the parameters of the
    effect (\code{After} has a delay and a message;
    \code{SetTitle} has a title string).
  \item \code{struct Batch \{ std::vector<Cmd> cmds; \}}: a
    \code{Batch} contains \emph{other} \code{Cmd}s, so the runtime
    can interpret a list of effects atomically. Note this is
    self-referential --- \code{Cmd} contains a vector of
    \code{Cmd} --- which works because \code{Batch}'s storage is
    in a \code{std::vector} (heap-allocated, no compile-time
    recursion).
  \item \code{struct Task \{ std::move\_only\_function<...> run;
    \}}: a Task carries a callable that the runtime will invoke on
    a background thread, passing in a callback the task can use to
    deliver messages back. We met
    \code{std::move\_only\_function} in Section
    \ref{sec:cpp:lambdas}; we use it here because the callable
    often captures move-only state (channels, file handles, etc.).
  \item \code{using Variant = std::variant<None, Quit, ...>}: the
    closed sum of every effect kind. This is a type alias inside
    the class; it gives the variant a stable name.
  \item \code{Cmd() : v\_(None\{\}) \{\}}: default constructor sets
    the variant to \code{None}, meaning ``no effect''. A default
    \code{Cmd} is harmless.
  \item Static factories \code{none()}, \code{quit()},
    \code{set\_title(...)}: convenience builders. Each returns a
    \code{Cmd} holding the right variant alternative. They are
    \code{[[nodiscard]]} so you cannot accidentally drop one.
  \item \code{template <typename Visitor> auto visit(Visitor\&\&
    vis) const}: the public interface for inspecting the effect.
    Forwards to \code{std::visit} on the inner variant.
    \code{Visitor\&\&} is a \emph{forwarding reference}; combined
    with \code{std::forward}, it preserves whether the caller
    passed an lvalue or rvalue. You can read this idiom as
    ``perfectly forward the visitor into \code{std::visit}''.
  \item \code{Variant v\_}: the actual storage. A single member
    variant; one effect at a time.
\end{itemize}

The runtime that consumes a \code{Cmd<Msg>} calls \code{visit} with
an \code{overload} matching every alternative. If a new effect is
added, the runtime visit is the one place that has to change ---
user code that builds a \code{Cmd} via the factories does not even
recompile (well, of course it does, but it does not need to be
edited).

If you understood that fragment, you have just read the core type
of the runtime layer. Chapter~\ref{ch:cmdsub} is the rest of the
story: which effects exist, what each does, and how the runtime
interprets them.

\section{Reading compiler errors}
\label{sec:cpp:errors}

\cpp{} compiler errors are notoriously long. A single misplaced
semicolon can produce a paragraph of diagnostics; a template error
can produce screens of them. Here is a survival guide.

\subsection{Read the first error first}

The rest are almost always consequences of the first. Fix the first
and recompile; many of the others usually evaporate.

\subsection{Look for ``required from'' or ``in instantiation of''}

Template errors print a long ``stack'' of

\begin{lstlisting}[style=shell]
foo.cpp:10:5:   in instantiation of 'X<int>' requested here
bar.hpp:42:7:   in instantiation of 'Y<X<int>>' requested here
baz.hpp:99:3:   required from here
qux.hpp:12:1:   error: no type named 'value_type' in 'X<int>'
\end{lstlisting}

The \emph{deepest} error is the actual reason; the lines above are
the path the compiler took to get there. Start at the bottom; work
up.

\subsection{Concepts errors are tractable}

If the failing template is constrained by a concept, the error names
the concept and the unmet requirement directly. This is one of the
killers reasons to use concepts. Compare:

\begin{lstlisting}[style=shell]
# without concepts:
error: no match for 'operator+' (operand types are 'File' and 'File')
... 50 lines of template instantiation context ...

# with concepts:
error: 'add(File, File)' does not satisfy concept 'Addable'
  note: the required expression '(a + b)' is invalid
\end{lstlisting}

When a maya API takes a constrained template parameter, the error
you get if you violate the constraint is a single, named, readable
line.

\subsection{Common shapes you will see}

\begin{itemize}[leftmargin=*]
  \item ``\code{no matching function for call to ...}'' --- you
    called a function with the wrong argument types. The compiler
    lists every candidate and says why each was rejected.
  \item ``\code{undefined reference to ...}'' --- a \emph{linker}
    error, not a compiler error. You declared a function but did
    not define it (or did not link the file that does).
  \item ``\code{expected ';' before ...}'' --- a missing semicolon
    somewhere \emph{earlier}, often at the end of a struct
    declaration.
  \item ``\code{use of deleted function ...}'' --- you tried to
    copy something that is move-only (e.g.\ a \code{unique\_ptr}).
    Add \code{std::move}.
  \item ``\code{conversion from X to Y is ambiguous}'' --- two
    different paths lead to the conversion. Disambiguate with an
    explicit \code{static\_cast}.
\end{itemize}

\begin{warning}
Compiler errors involving \code{std::tuple}, \code{std::variant}, or
\code{std::function} can run to thousands of characters. Pipe to
\code{less} or use an editor with line-wrap. The relevant signal is
still the first error.
\end{warning}

\begin{recap}
The \cpp{} \maya{} uses, in one paragraph: templates with type and
non-type parameters; \code{constexpr}/\code{consteval} for compile-
time evaluation; concepts and \code{requires} for named constraints;
\code{std::variant} + \code{std::visit} + \code{overload} for closed
sum types; \code{std::expected} for structured errors; lambdas with
explicit capture and \code{std::move\_only\_function} for storage;
move semantics to avoid copies; RAII to make resources self-managing.
Together they support one discipline: encode invariants in the type
system, make impossible states unrepresentable, defer side effects to
the edge.
\end{recap}

\section*{Workshop}

The exercises here are calibrated for someone who has never used the
feature. If a step takes you fifteen minutes and three Google
searches, that is the right speed.

\begin{enumerate}[leftmargin=*]
  \item \textbf{RAII guard.} Write a class \code{Timer} whose
    constructor records \code{std::chrono::steady\_clock::now()} and
    whose destructor prints the elapsed time. Use it at the top of
    a function; verify the message appears on return.
  \item \textbf{Templates and NTTPs.} Implement a \code{FixedArray<T,
    N>} class template with a member \code{T data[N]}, a
    \code{size()} method returning \code{N}, and \code{operator[]}.
    Instantiate \code{FixedArray<int, 4>} and \code{FixedArray<int,
    8>}; convince yourself they are different types by trying to
    assign one to the other.
  \item \textbf{NTTP of structural type.} Type out the \code{Str<N>}
    template from this chapter. Define a class template \code{Tag<Str
    S>} parameterised by it. Instantiate \code{Tag<"hello">} and
    \code{Tag<"world">} and call a method that returns the string.
  \item \textbf{Concept.} Write a concept \code{Drawable} requiring
    a \code{void draw() const} method. Use it on a template function
    \code{render(const T\& t)}. Make two structs satisfying it, one
    not, and observe the error.
  \item \textbf{Variant.} Define
    \code{using Shape = std::variant<Circle, Square, Triangle>}
    with three trivial structs (each with one field). Write a
    \code{double area(const Shape\&)} using \code{std::visit} and the
    \code{overload} helper.
  \item \textbf{Expected.} Write
    \code{std::expected<int, std::string> safe\_div(int a, int b)}.
    Call it with \code{(10, 2)} and \code{(10, 0)}; print each
    result.
  \item \textbf{Lambda with move capture.} Build a
    \code{std::vector<int>}, capture it into a lambda by move, and
    return the lambda from a function. Call the returned lambda;
    confirm the vector outlives the function it came from.
\end{enumerate}
