\chapter{The compile-time DSL}
\label{ch:dsl}

\maya{}'s most distinctive feature is its DSL --- a tiny embedded
language inside \cpp{} for describing UI trees. The DSL is unusual
because it lives entirely at the type level: an expression like

\begin{lstlisting}
t<"Hello"> | Bold | Fg<100, 180, 255>
\end{lstlisting}

is a \emph{single value} whose type encodes every choice you made
(\emph{the string}, \emph{bold}, \emph{the colour}). This chapter
shows how that works, why the compiler refuses certain expressions,
and how to escape into the runtime when you need to.

\begin{aside}
If you have written HTML or React, you have written a DSL: a small,
specialised vocabulary for the one thing you are doing (describing
UI). \maya{}'s DSL is the same idea, but it sits inside \cpp{}, runs
at compile time, and refuses to compile if you build an illegal
tree. The bottom-up explanation in this chapter is long, but the
payoff is being able to read any expression in any \maya{} program
and know what every token contributes to the type.
\end{aside}

\section{From scratch: what a DSL even is}
\label{sec:dsl:what}

The phrase ``Domain-Specific Language'' has two flavours in
programming.

An \textbf{external DSL} is a language with its own parser, lexer,
and often its own interpreter: SQL, regular expressions, JSON, CSS,
the shell. You write it in a different syntax from the host
language, and a separate engine reads it.

An \textbf{embedded} (or \emph{internal}) DSL is a sub-vocabulary
within the host language. It looks like normal code, but the
operators and identifiers have been overloaded to express a different
domain. JavaScript's \code{describe("thing", () => \dots)} from
Mocha is an embedded DSL for testing. \cpp{}'s \code{std::cout <<
thing} is the world's most ubiquitous embedded DSL --- the \code{<<}
operator does \emph{stream insertion} when applied to streams, not
bit-shift.

The \maya{} DSL is the embedded kind. Every piece of it is legal
\cpp{}; the magic is in operator overloads, variable templates, and
the fact that we have made the result types interesting enough that
the compiler can reason about them.

Four properties define a good embedded DSL:

\begin{enumerate}[leftmargin=*]
  \item \textbf{It reads like the thing it describes.}
    \code{v(t<"a">, t<"b">)} looks vertical because we picked
    \code{v} for that role.
  \item \textbf{It composes with the rest of the language.} You can
    pass a DSL fragment to a function, return one from a function,
    store one in a variable.
  \item \textbf{The host's type checker still works.} A mistake in
    the DSL is a host-language error, not a runtime ``DSL
    interpreter blew up''.
  \item \textbf{There is an escape hatch back to the host
    language.} \maya{}'s \code{dyn([]{...})} is exactly this.
\end{enumerate}

In the rest of the chapter we will see how \maya{} achieves all four.

\section{Non-type template parameters, properly}
\label{sec:dsl:nttp}

The foundation of the entire DSL is the \cpp{}20 / 26 ability to use
structured values as template parameters. We touched on this in
Chapter~\ref{ch:cpp}; here is the longer version, because the DSL
leans on it constantly.

A \textbf{template parameter} can be a type (\code{typename T}), but
it can also be a \emph{value}:

\begin{lstlisting}
template <int N> struct Array { int data[N]; };
Array<8>  a;   // N = 8
Array<16> b;   // N = 16, different type from a
\end{lstlisting}

The parameter \code{N} is an integer, not a type. The template is
instantiated once per distinct value of \code{N}, and each
instantiation is a separate type. \code{Array<8>} and \code{Array<16>}
do not interconvert.

\subsection{Why integers are not enough}

Integers, pointers, and enum values have always been allowed as
non-type template parameters (NTTPs). What you could \emph{not} do,
until \cpp{}20, was use a class type as an NTTP. There was no way to
write

\begin{lstlisting}
struct Point { int x; int y; };
template <Point P> struct Thing { /* ... */ };
Thing<Point{3, 4}> t;     // illegal in C++17
\end{lstlisting}

This matters for us because a UI description involves not just
numbers but \emph{compound data}: a string is a class type, a style
is a class type, a colour is three bytes. If we cannot put those in
template parameters, the DSL is severely limited.

\cpp{}20 lifted the restriction for \textbf{structural} class types:
any literal type whose members are all public, with no user-defined
oddities, can be an NTTP. \cpp{}26 extended it further. The
consequence is that \code{maya::dsl::Str}, \code{maya::dsl::CTStyle},
and \code{maya::dsl::BoxCfg} can all appear as template parameters,
and the DSL works.

\subsection{\code{Str}: the building block}

The simplest structural NTTP in \maya{} is \code{Str} ---
a fixed-size string class:

\begin{lstlisting}
template <std::size_t N>
struct Str {
    char data[N]{};
    consteval Str(const char (&s)[N]) {
        for (std::size_t i = 0; i < N; ++i) data[i] = s[i];
    }
    consteval std::size_t size() const { return N - 1; }
    constexpr operator std::string_view() const { return {data, N - 1}; }
};
\end{lstlisting}

Line by line:

\begin{itemize}[leftmargin=*]
  \item \code{template <std::size\_t N> struct Str}: parameterised
    by the length \code{N}.
  \item \code{char data[N]\{\}}: a fixed array of \code{N}
    characters, zero-initialised.
  \item \code{consteval Str(const char (\&s)[N])}: constructor that
    \emph{must} run at compile time, taking a reference to a
    \code{char} literal of length \code{N}. The compiler deduces
    \code{N} from the literal's size.
  \item The loop copies the literal into \code{data}.
\end{itemize}

Now you can write \code{Str{"Hello"}} as a constant expression. The
resulting value has a \emph{type} that depends on the length and a
\emph{value} that contains the bytes --- and crucially, both are
available to other templates.

\subsection{Putting \code{Str} into the type system}

\begin{lstlisting}
template <Str S>
struct TextNode {
    static constexpr std::string_view text() { return {S.data, S.size()}; }
};

TextNode<"Hello"> a;
TextNode<"World"> b;
// decltype(a) and decltype(b) are different types,
// even though both store no data per-instance.
\end{lstlisting}

\code{TextNode<"Hello">} and \code{TextNode<"World">} are distinct
types. The string contents are part of the \emph{type identity},
not a field. Two instances of \code{TextNode<"Hello">} occupy zero
bytes and are interchangeable.

Why does this matter? Because the compiler can now reason about your
strings: it can choose overloads based on which string was used, it
can refuse to mix incompatible strings, and it can fold the string's
length into compile-time arithmetic --- without you writing any of
that by hand.

\subsection{\code{CTStyle}: a structural style}

The second key NTTP type:

\begin{lstlisting}
struct CTStyle {
    bool has_fg = false, has_bg = false;
    uint8_t fg_r = 0, fg_g = 0, fg_b = 0;
    uint8_t bg_r = 0, bg_g = 0, bg_b = 0;
    bool bold_ = false, dim_ = false, italic_ = false;
    bool underline_ = false, strike_ = false, inverse_ = false;

    consteval CTStyle merge(CTStyle o) const { /* combine fields */ }
    [[nodiscard]] Style runtime() const       { /* materialise */ }
};
\end{lstlisting}

A \code{CTStyle} value is a \textbf{bag of compile-time style
flags}: bold yes/no, foreground colour yes/no plus three RGB bytes,
and so on. The struct is structural (all-public fields, no virtuals,
no custom comparison), so it qualifies as an NTTP. The
\code{merge} method is \code{consteval} --- it can be called inside
template arguments to compute new \code{CTStyle}s from old ones.

A style tag exists for every bool field:

\begin{lstlisting}
template <CTStyle V>
struct StyTag {
    static constexpr CTStyle value = V;
};

inline constexpr StyTag<CTStyle{.bold_ = true}>      Bold{};
inline constexpr StyTag<CTStyle{.dim_ = true}>       Dim{};
inline constexpr StyTag<CTStyle{.italic_ = true}>    Italic{};
inline constexpr StyTag<CTStyle{.underline_ = true}> Underline{};
\end{lstlisting}

\code{Bold} is a \code{StyTag} whose template argument is a
\code{CTStyle} with one field set. The variable \code{Bold} itself
is a zero-size instance --- the \code{CTStyle} lives in the type.

\subsection{\code{Fg<R,G,B>}: colours in the type system}

Colours are encoded the same way: a \code{CTStyle} with the
foreground fields set, exposed through a variable template:

\begin{lstlisting}
namespace detail_rgb {
    template <uint8_t R, uint8_t G, uint8_t B>
    consteval CTStyle make_fg() {
        return CTStyle{.has_fg=true, .fg_r=R, .fg_g=G, .fg_b=B};
    }
}

template <uint8_t R, uint8_t G, uint8_t B>
inline constexpr StyTag<detail_rgb::make_fg<R, G, B>()> Fg{};
\end{lstlisting}

\code{Fg<100, 180, 255>} is a zero-size value whose type is a
\code{StyTag} whose template parameter is a \code{CTStyle} whose
fields are exactly the RGB you asked for. The numbers fed in become
part of the type. Nothing about colour leaks into runtime data
until \code{.build()}.

\begin{funfact}
The MSVC compiler had (and partly still has) a long-standing bug
around \code{constexpr} evaluation of designated-initialised NTTPs.
If you look in \file{dsl.hpp} you will see a small namespace
\code{detail\_rgb} with \code{make\_fg} / \code{make\_bg} helpers ---
that existence-of-helpers is the workaround. The compiler bug ticket
is cited in a comment. Real libraries are full of small monuments
like this; each is a story of someone who hit the same wall before
you.
\end{funfact}

\begin{idea}
NTTPs let you put \emph{values} into the type system. \maya{} uses
that to encode strings (\code{Str}), styles (\code{CTStyle}),
colours (RGB triples), borders, paddings, and growth factors. Once a
value lives in a type, the compiler can check rules between values
at compile time --- which is the whole point.
\end{idea}

\section{The shape}

The DSL has six kinds of node:

\begin{center}
\small
\begin{tabular}{ll}
\code{TextNode<Str, CTStyle>}     & compile-time text \\
\code{RuntimeTextNode<S>}         & runtime text (from a string or number) \\
\code{BoxNode<Dir, Cfg, Cs...>}   & vertical or horizontal container \\
\code{DynNode<F>}                 & lambda escape hatch \\
\code{MapNode<R, P>}              & project a range to elements \\
\code{SpacerNode}                 & a flexible gap \\
\end{tabular}
\end{center}

Every node satisfies the \code{Node} concept:

\begin{lstlisting}
template <typename T>
concept Node = requires(const T& n) {
    { n.build() } -> std::convertible_to<Element>;
};
\end{lstlisting}

A \code{Node} is anything with a \code{.build()} that yields a runtime
\code{Element}. That is the only contract; the rest is how nodes
combine.

\section{\code{t<"Hello">} --- string in a type}

The simplest node is the compile-time text node:

\begin{lstlisting}
template <Str S, CTStyle Sty = CTStyle{}>
struct TextNode {
    static constexpr Str  str   = S;
    static constexpr CTStyle style = Sty;

    Element build() const {
        return TextElement{std::string{S.view()}, Sty.runtime()};
    }
    operator Element() const { return build(); }
};

template <Str S>
inline constexpr TextNode<S> t{};
\end{lstlisting}

The template parameters are: an \code{Str} (the string contents, an
NTTP from Chapter~\ref{ch:cpp}) and a \code{CTStyle} (a structural
type holding the compile-time style). \code{t<"Hello">} is the
variable template: \code{TextNode<"Hello", CTStyle\{\}>\{\}}.

Two facts to internalise:
\begin{itemize}[leftmargin=*]
  \item \code{t<"a">} and \code{t<"b">} have \emph{different types}.
    Their string contents live in the type system, not in a field.
  \item \code{TextNode} has no nonstatic data members. It is a zero-size
    type. Two instances of \code{TextNode<"a">} are
    indistinguishable; the compiler treats them as the same value.
\end{itemize}

\section{Pipes are operator overloads}

The DSL's pipe (\code{|}) is just \code{operator|}. \maya{} defines
overloads that match \code{(Node, StyTag)} and return a new node with
the style merged in:

\begin{lstlisting}
template <Str S, CTStyle Sty, CTStyle V>
constexpr auto operator|(TextNode<S, Sty>, StyTag<V>) {
    return TextNode<S, Sty.merge(V)>{};   // new type, same instance shape
}
\end{lstlisting}

Read the return type: the merged style is computed by \code{Sty.merge(V)},
which is \code{constexpr}, so the result is a brand-new
\code{TextNode<S, Sty'>}. Apply two style tags and you get yet another
type. The pipeline is sequential type rewriting at compile time.

\begin{idea}
Because the result of each pipe is a new type, the compiler can
enforce \emph{type-state} rules: a method is only available on a node
that has reached a particular state.
\end{idea}

This is the trick behind the border colour example. The pipe overload
for \code{BColTag} (a border colour tag) has a \code{requires} clause:

\begin{lstlisting}
template <FlexDirection Dir, BoxCfg Cfg, typename... Cs,
          uint8_t R, uint8_t G, uint8_t B>
    requires (Cfg.has_border)
constexpr auto operator|(BoxNode<Dir, Cfg, Cs...> n, BColTag<R, G, B>) {
    /* return a new BoxNode with border colour set */
}
\end{lstlisting}

If you call \code{v(...) | bcol<60, 65, 80>} without first applying
\code{border\_<Round>}, \code{Cfg.has\_border} is \code{false}, the
\code{requires} clause fails, no overload matches, and the compiler
prints an error. It is the same mechanism a Rust crate would use with
typestate traits.

\section{\code{v()} and \code{h()} --- variadic containers}

A box takes any number of children and a direction:

\begin{lstlisting}
template <FlexDirection Dir, BoxCfg Cfg, typename... Cs>
struct BoxNode {
    std::tuple<Cs...> children;

    Element build() const {
        BoxBuilder b{Dir, Cfg};
        std::apply([&](const auto&... cs) {
            (b.add(cs.build()), ...);   // fold-expression over children
        }, children);
        return b.finalize();
    }
};

template <typename... Cs>
constexpr auto v(Cs... cs) {
    return BoxNode<FlexDirection::Column, BoxCfg{}, Cs...>{
        std::tuple{std::move(cs)...}
    };
}

template <typename... Cs>
constexpr auto h(Cs... cs) {
    return BoxNode<FlexDirection::Row, BoxCfg{}, Cs...>{
        std::tuple{std::move(cs)...}
    };
}
\end{lstlisting}

Two \cpp{} features worth pausing on:

\begin{itemize}[leftmargin=*]
  \item \textbf{Variadic templates} (\code{typename... Cs}): you can
    declare ``zero or more types''. \code{Cs...} is a \emph{parameter
    pack}; you expand it with \code{...}.
  \item \textbf{Fold expressions} (\code{(b.add(cs.build()), ...)}):
    short for ``apply this expression to every element of the
    pack''. The comma is the \emph{fold operator} --- evaluate each,
    discard intermediate results.
\end{itemize}

The result: \code{v(t<"a">, t<"b">, t<"c">)} is a single
\code{BoxNode<Column, BoxCfg\{\}, TextNode<"a">, TextNode<"b">,
TextNode<"c">>}, holding a \code{tuple} of three zero-size text
nodes. Total runtime cost: zero bytes (the tuple of empty types is
empty).

\section{\code{BoxCfg} --- the type-state}

\code{BoxCfg} is a \code{constexpr} struct that records every layout
decision applied to a box:

\begin{lstlisting}
struct BoxCfg {
    int pad_top    = 0, pad_right = 0, pad_bottom = 0, pad_left = 0;
    int gap        = 0;
    int grow_num   = 0;
    bool has_border       = false;
    BorderStyle border    = BorderStyle::None;
    bool has_border_color = false;
    Color border_color{};
    Align align_items    = Align::Auto;
    Justify justify      = Justify::Auto;
    // ... etc.

    constexpr BoxCfg with_padding(int t, int r, int b, int l) const {
        BoxCfg c = *this;
        c.pad_top = t; c.pad_right = r; c.pad_bottom = b; c.pad_left = l;
        return c;
    }
    constexpr BoxCfg with_border(BorderStyle bs) const {
        BoxCfg c = *this; c.has_border = true; c.border = bs; return c;
    }
    // ...
};
\end{lstlisting}

Every layout pipe (\code{pad<1>}, \code{gap\_<1>}, \code{border\_<Round>},
\code{grow\_<2>}) returns a new \code{BoxNode} whose \code{BoxCfg}
template parameter has been replaced via the corresponding
\code{with\_*} method. The type carries the full configuration; the
final \code{.build()} reads the type and produces the runtime
\code{BoxElement}.

\section{Static, dynamic, and the bridge}

A pure-compile-time tree is fine for a banner but useless for a real
app: the count, the file list, the current time are all known only at
runtime. \maya{} provides three bridge nodes.

\subsection{\code{text(...)} --- runtime strings}

\begin{lstlisting}
text("hello")                // wraps a string_view
text(std::to_string(n))      // wraps a string
text(42)                     // wraps an int, formatted at build time
text(3.14)                   // wraps a double
\end{lstlisting}

\code{text} returns a \code{RuntimeTextNode<S>} --- a node whose
content is a runtime value but which still supports the DSL pipes.
Its \code{build()} formats the value and produces a
\code{TextElement}.

\subsection{\code{dyn([](){ ... })} --- arbitrary lambdas}

\begin{lstlisting}
dyn([&] {
    return loading ? text("Loading...") | Dim
                   : text("Ready")     | Bold | Fg<80, 220, 120>;
})
\end{lstlisting}

\code{dyn} wraps a lambda that returns anything convertible to
\code{Element}. \code{build()} just calls the lambda. This is the
universal escape hatch: when you cannot express something in the DSL
itself, write a lambda.

\subsection{\code{map(range, projection)} --- list comprehension}

\begin{lstlisting}
std::vector<File> files = ...;

auto list = v(
    t<"Files:"> | Bold,
    map(files, [](const File& f) {
        return h(text(f.name) | Bold, space, text(f.size) | Dim);
    })
);
\end{lstlisting}

\code{map} produces a \code{MapNode<R, P>}. At \code{build()} time it
walks the range, applies the projection, and produces an
\code{ElementList} that is spliced into the parent's children.

\section{Reading a real pipeline}

Put it all together. Here is a typical maya panel:

\begin{lstlisting}
auto panel = v(
    t<"Dashboard"> | Bold | Fg<100, 180, 255>,
    sep,
    dyn([&] { return text("Active users: " + std::to_string(state.users)); }),
    map(state.items, [](const auto& it) {
        return h(text(it.name) | Bold, space, text(it.value) | Dim);
    }),
    space,
    t<"[q] quit"> | Dim
) | border_<Round> | bcol<60, 65, 80> | pad<1, 2> | gap_<1>;
\end{lstlisting}

What is the type of \code{panel}? Approximately:

\begin{lstlisting}
BoxNode<
    FlexDirection::Column,
    BoxCfg{ /* round border, colour, pad 1/2, gap 1 */ },
    TextNode<"Dashboard", { bold, fg=(100,180,255) }>,
    SepNode,
    DynNode<lambda1>,
    MapNode<vector<Item>, lambda2>,
    SpacerNode,
    TextNode<"[q] quit", { dim }>
>
\end{lstlisting}

The compiler holds every choice you made --- the strings, the styles,
the border, the padding --- in this one large type. \code{panel.build()}
then walks it once, recursively, and produces an \code{Element} tree
that the layout engine and renderer take from there.

\section{Why compile-time?}

You could write this DSL with runtime polymorphism (each node a
\code{shared\_ptr<Node>}, each pipe a method that mutates fields). It
would work. But:

\begin{itemize}[leftmargin=*]
  \item \textbf{Allocations vanish.} A purely compile-time tree of
    zero-size types has no heap traffic.
  \item \textbf{Errors are at compile time.} Border colour without a
    border, negative padding (template parameter constraint
    \code{N >= 0}), wrong type of child --- all rejected before the
    program runs.
  \item \textbf{The compiler can fold.} A \code{constexpr} tree can
    be assigned to a \code{constexpr} variable and rendered at startup
    with no parsing.
\end{itemize}

The trade-off: error messages can be long. The first time you see a
500-line template error, breathe. The bottom of the message --- the
\code{required from here} line and the constraint that failed --- is
usually the actionable bit.

\section{Where the runtime takes over}

\code{.build()} is the boundary. Above it: types, NTTPs, fold
expressions, the world the compiler sees. Below it: \code{Element}, a
runtime value the layout and renderer understand. The next chapter
goes through that runtime side.

\begin{recap}
\maya{}'s DSL is a set of node types that compose with \code{|} into
larger node types. Strings, styles, borders, paddings, and tree
structure all live as template parameters. Type-state constraints
make invalid trees uncompilable. Runtime content enters through
\code{text}, \code{dyn}, and \code{map}. \code{.build()} converts the
compile-time tree into a runtime \code{Element}.
\end{recap}

\section{Every pipe, named and explained}
\label{sec:dsl:pipes-tour}

The DSL has roughly twenty overloads of \code{operator|}. Reading
the source for each takes ten seconds; reading them together is
useful, because the table makes it obvious which kinds of node
accept which kinds of modifier. Here is the full list, grouped by
the left-hand side.

\subsection{Pipes that take a \code{TextNode}}

\begin{center}
\small
\begin{tabular}{lll}
\textbf{Pipe} & \textbf{Effect} & \textbf{Result type} \\
\hline
\code{TextNode | StyTag<V>}    & merge style V       & \code{TextNode<S, Sty.merge(V)>} \\
\code{TextNode | WidthTag<W>}  & wrap in fixed box   & \code{BoxNode<Row, ct\_width=W>} \\
\code{TextNode | HeightTag<H>} & wrap in fixed box   & \code{BoxNode<Row, ct\_height=H>} \\
\end{tabular}
\end{center}

Text nodes accept any style modifier and nothing else. Adding
\code{pad<1>} to a bare text node is a compile error --- padding is
a container property; text has no notion of inset space.

\subsection{Pipes that take a \code{RuntimeTextNode}}

The runtime text node (built by \code{text(...)}) gets the same
style pipes plus two wrap-mode pipes:

\begin{center}
\small
\begin{tabular}{ll}
\textbf{Pipe} & \textbf{Effect} \\
\hline
\code{| StyTag<V>}    & merge style into the runtime node \\
\code{| clip}         & truncate with ellipsis on overflow \\
\code{| nowrap}       & never break the line, may overflow \\
\end{tabular}
\end{center}

The wrap-mode pipes do not exist on compile-time \code{TextNode}
because compile-time text is meant to be short and known; the
ellipsis case only matters when you wrap a runtime value of unknown
length.

\subsection{Pipes that take a \code{BoxNode}}

This is where the real type-state machine lives. A box accepts every
kind of modifier:

\begin{center}
\small
\begin{tabular}{lll}
\textbf{Pipe} & \textbf{Effect} & \textbf{Notes} \\
\hline
\code{| StyTag<V>}        & merge style on the box itself      & all boxes \\
\code{| PadTag<T,R,B,L>}  & set inner padding                  & all boxes \\
\code{| GapTag<G>}        & set gap between children           & all boxes \\
\code{| BorderTag<BS>}    & set border style                   & enables \code{bcol} pipe \\
\code{| BColTag<R,G,B>}   & set border colour                  & \textbf{requires} \code{has\_border} \\
\code{| GrowTag<G>}       & set flex grow factor               & all boxes \\
\code{| WidthTag<W>}      & set fixed width                    & all boxes \\
\code{| HeightTag<H>}     & set fixed height                   & all boxes \\
\end{tabular}
\end{center}

Four of the modifier types have a \code{static\_assert} on the
template parameter to reject obvious nonsense at compile time:

\begin{itemize}[leftmargin=*]
  \item \code{PadTag}: every value must be \code{>= 0}. Negative
    padding is a category error.
  \item \code{GapTag}: \code{G >= 0}.
  \item \code{GrowTag}: \code{G >= 0}.
  \item \code{WidthTag}, \code{HeightTag}: \code{> 0}. A width of
    zero is a hidden box, which we have a different way to express
    (\code{visible(false)}).
\end{itemize}

The one with the named requires-clause is \code{BColTag}:

\begin{lstlisting}
template <FlexDirection Dir, BoxCfg Cfg, typename... Cs,
          uint8_t R, uint8_t G, uint8_t B>
    requires (Cfg.has_border)
constexpr auto operator|(BoxNode<Dir, Cfg, Cs...> n, BColTag<R, G, B>);
\end{lstlisting}

The constraint \code{requires (Cfg.has\_border)} forbids the
overload when the box has no border yet. We will look at the error
message in section~\ref{sec:dsl:errors}.

\subsection{The style-merging pipe}

A quietly powerful one: two style tags combine into one.

\begin{lstlisting}
template <CTStyle A, CTStyle B>
constexpr auto operator|(StyTag<A>, StyTag<B>) {
    return StyTag<A.merge(B)>{};
}
\end{lstlisting}

This is the foundation of building reusable style presets:

\begin{lstlisting}
constexpr auto heading = Bold | fg<0xFFFFFF>;
constexpr auto error   = Bold | fg<0xFF4444>;
constexpr auto subtle  = Dim  | Italic;

t<"Title">     | heading
text(msg)      | error
t<"caption">   | subtle
\end{lstlisting}

The combined tag is itself a \code{StyTag<...>}, so it composes
further: \code{heading | Underline} is legal. Compose your design
system this way and applying it is a single token.

\subsection{The hex-colour shorthand}

\code{Fg<R, G, B>} is verbose for designers, who think in hex. The
DSL provides \code{fg<Hex>} and \code{bg<Hex>}:

\begin{lstlisting}
t<"warn">  | fg<0xFFAA00>
t<"error"> | fg<0xFF4444>
v(...)     | bg<0x1A1A2E>
\end{lstlisting}

The \code{consteval} helper that decodes the hex into RGB also
validates it: \code{fg<0xFFFFFFFF>} (32 bits) throws a string
literal from inside \code{consteval}, which the compiler reports as
a constant-expression failure --- ``\code{fg<>: hex value exceeds
0xFFFFFF}''. Compile-time error, not a silent black box.

\section{The type-state machine, drawn out}
\label{sec:dsl:state-machine}

The phrase \textbf{type-state} comes from research on programming
languages: encode the legal sequence of operations on an object into
its type, so that calling a method at the wrong stage is a type
error. Rust's \code{Builder} idioms are typical: the constructor
returns a \code{Builder<Empty>}, the \code{name(\dots)} method
returns a \code{Builder<Named>}, and \code{build()} is only defined
on \code{Builder<Named>}.

\maya{}'s \code{BoxNode} works the same way, with \code{BoxCfg} as
the state.

\paragraph{Initial state.} \code{v(\dots)} returns a
\code{BoxNode<Column, BoxCfg\{\}, \dots>}: every flag in
\code{BoxCfg} is false / zero. From this state:

\begin{itemize}[leftmargin=*]
  \item Padding, gap, grow, width, height, raw style: all available.
    Each returns a new \code{BoxNode} whose \code{BoxCfg} has the
    corresponding field updated.
  \item Border: available. Returns a \code{BoxNode} with
    \code{has\_border = true}.
  \item Border colour: \emph{not} available; the requires-clause
    fails because \code{has\_border} is false.
\end{itemize}

\paragraph{After \code{| border\_<Round>}.} \code{has\_border}
becomes true. Border colour becomes available.

\paragraph{Drawn as a state diagram.}

\begin{center}
\small
\begin{tabular}{rcl}
 \code{BoxNode<Cfg(has\_border=false)>}
  & $\xrightarrow{\;\;|\;\code{border\_<R>}\;\;}$
  & \code{BoxNode<Cfg(has\_border=true)>} \\
 \code{BoxNode<Cfg(has\_border=false)>}
  & $\xrightarrow{\;\;|\;\code{bcol<\dots>}\;\;}$
  & \textsc{compile error} \\
 \code{BoxNode<Cfg(has\_border=true)>}
  & $\xrightarrow{\;\;|\;\code{bcol<\dots>}\;\;}$
  & \code{BoxNode<\dots, has\_border\_color=true>} \\
\end{tabular}
\end{center}

The state machine is tiny --- two states for the border flag --- but
the pattern scales. Any rule of the form ``\emph{X} is only legal
after \emph{Y}'' can be encoded the same way, by toggling a field in
the \code{BoxCfg} on the \emph{Y} operation and adding a
\code{requires (Cfg.\emph{field})} to the \emph{X} overload.

\begin{idea}
Type-state in a DSL means: the legal sequence of pipes is encoded
into the chain of types. The compiler refuses pipes the current type
does not permit. The error has a name (a \code{requires} clause) and
a single source line.
\end{idea}

\section{Decoding the compile errors}
\label{sec:dsl:errors}

A powerful type system gets you powerful errors. The first time you
see a 200-line template error, the impulse is to scroll up and
panic; the trick is knowing where to look.

The shape of every constraint-failure error from the DSL is:

\begin{enumerate}[leftmargin=*]
  \item A line saying \emph{no viable overload of \code{operator|}}.
  \item One or more notes saying \emph{candidate template ignored:
    constraints not satisfied}.
  \item For each candidate, the constraint that did not hold:
    \code{Cfg.has\_border} was \code{false}, \code{T >= 0} did not
    hold for \code{T = -1}, and so on.
  \item Finally, the line in \emph{your} code that triggered it.
\end{enumerate}

\subsection{The border-colour error}

The source mistake:

\begin{lstlisting}
auto ui = v(t<"hi">) | bcol<60, 65, 80>;     // forgot border first
\end{lstlisting}

The heart of the error from a modern \code{g++}:

\begin{lstlisting}[style=shell]
error: no match for 'operator|' (operand types are
    'maya::dsl::BoxNode<maya::FlexDirection::Column,
      maya::dsl::BoxCfg{...has_border=false...}, ...>'
    and 'const maya::dsl::BColTag<60, 65, 80>')
note: candidate: 'template<...> constexpr auto
    maya::dsl::operator|(BoxNode<Dir, Cfg, Cs...>, BColTag<R, G, B>)
        requires (Cfg.has_border)'
note: constraints not satisfied
note:   the expression 'Cfg.has_border' evaluated to 'false'
\end{lstlisting}

The relevant fact is in the last note: \code{Cfg.has\_border}
evaluated to false. Fix:

\begin{lstlisting}
auto ui = v(t<"hi">) | border_<Round> | bcol<60, 65, 80>;
\end{lstlisting}

\subsection{The negative-padding error}

The source mistake:

\begin{lstlisting}
auto ui = v(t<"hi">) | pad<-1>;
\end{lstlisting}

The error:

\begin{lstlisting}[style=shell]
error: static assertion failed: Padding values must be non-negative
  in instantiation of 'struct maya::dsl::PadTag<-1, -1, -1, -1>'
\end{lstlisting}

A \code{static\_assert} caught it before any pipe overload was
even considered. The error names exactly what is wrong.

\subsection{The wrong-child-type error}

The DSL's \code{v} and \code{h} take children that satisfy the
\code{DslChild} concept --- either a \code{Node} (anything with
\code{.build()}) or a \code{std::ranges::range} of \code{Element}.
Passing something else fails the concept check:

\begin{lstlisting}
auto bad = v(42);                       // int is not a DslChild
\end{lstlisting}

The error roughly:

\begin{lstlisting}[style=shell]
error: no matching function for call to 'v(int)'
note: candidate template ignored: constraints not satisfied [with Cs = {int}]
note: because 'int' does not satisfy 'DslChild'
\end{lstlisting}

The fix is to wrap the value in \code{text(\dots)} or its appropriate
constructor: \code{v(text(42))}.

\subsection{Strategy for reading a template error}

Three rules that hold for every \maya{} compile failure:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Read the bottom first.} The line that says
    \emph{required from here} or \emph{in the expansion of \dots}
    is closer to your source than the wall of template-instantiation
    backtrace above it.
  \item \textbf{Look for ``constraints not satisfied'' / ``static
    assertion failed''.} Those are the human-readable handles. The
    text after them names the rule that broke.
  \item \textbf{Identify the offending NTTP value.} The compiler
    will print the \code{BoxCfg} or \code{CTStyle} with its fields;
    the field whose value is wrong (\code{has\_border=false},
    \code{pad\_t=-1}) is the actionable bit.
\end{enumerate}

Most of the wall of text is irrelevant; you are looking for two or
three lines.

\begin{warning}
Do \emph{not} \code{\#include <maya/maya.hpp>} into a translation
unit that you then refuse to use because the error is intimidating.
The errors are intimidating because they are precise. The fix is
usually one token away; scrolling past the error makes finding it
harder.
\end{warning}

\section{What \code{.build()} actually does}
\label{sec:dsl:build}

The boundary between compile time and runtime is the \code{.build()}
method. Every node provides it; the implementation depends on the
kind of node.

\paragraph{\code{TextNode<S, Sty>::build()}.}

\begin{lstlisting}
Element build() const {
    return Element{TextElement{
        .content = std::string{std::string_view{S}},
        .style   = Sty.runtime(),
    }};
}
\end{lstlisting}

The compile-time string \code{S} and style \code{Sty} are read out
of the type. \code{S} becomes a \code{std::string} (heap-allocated
if the small-string optimisation does not apply; for short banners
it usually does). \code{Sty.runtime()} converts the
\code{CTStyle} bag into a runtime \code{Style} object (Chapter~\ref{ch:elements}).
The result is a \code{TextElement} wrapped in the \code{Element}
variant.

\paragraph{\code{RuntimeTextNode<S>::build()}.}

\begin{lstlisting}
Element build() const {
    if constexpr (std::integral<S> && !std::same_as<S, bool>) {
        // format integer via std::to_chars
    } else if constexpr (std::floating_point<S>) {
        // format float via std::to_chars, 2 fractional digits
    } else {
        // S is a string-like: copy into a std::string
    }
}
\end{lstlisting}

The \code{if constexpr} picks the right path at compile time, so
there is no per-call branching. \code{text(42)} formats with no
heap allocation (uses a 32-byte stack buffer); \code{text("hi")}
allocates a \code{std::string}.

\paragraph{\code{BoxNode<Dir, Cfg, Cs...>::build()}.} The most
involved:

\begin{lstlisting}
Element build() const {
    return std::apply([](const auto&... cs) {
        auto b = maya::detail::box().direction(Dir);
        if constexpr (Cfg.pad_t || Cfg.pad_r || ...)
            b = std::move(b).padding(Cfg.pad_t, ...);
        if constexpr (Cfg.gap > 0)
            b = std::move(b).gap(Cfg.gap);
        if constexpr (Cfg.has_border)
            b = std::move(b).border(Cfg.bstyle);
        if constexpr (Cfg.has_border_color)
            b = std::move(b).border_color(Color::rgb(Cfg.bc_r, Cfg.bc_g, Cfg.bc_b));
        // ... style, dimensions, etc.
        return b(cs.build()...);   // fold-expression: build each child
    }, children);
}
\end{lstlisting}

Notice the \code{if constexpr} on every \code{BoxCfg} field. A box
with no border generates \emph{no} call to \code{.border()}; the
branch does not exist in the resulting binary. The compiler
specialises one \code{build()} per distinct \code{BoxCfg} and
throws out everything irrelevant.

The trailing \code{b(cs.build()...)} expands the children pack:
every child's \code{.build()} is invoked, and the resulting
\code{Element}s are handed to the runtime box builder via its
function-call operator (Chapter~\ref{ch:elements}).

\paragraph{\code{DynNode<F>::build()}.}

\begin{lstlisting}
Element build() const { return fn(); }
\end{lstlisting}

Literal one-liner. The lambda is invoked; whatever it returns is
converted to \code{Element}.

\paragraph{\code{MapNode<R, Proj>::build()}.}

\begin{lstlisting}
Element build() const {
    std::vector<Element> items;
    if constexpr (std::ranges::sized_range<R>) items.reserve(range.size());
    for (auto&& val : range) items.emplace_back(proj(val).build());
    return Element{ElementList{std::move(items)}};
}
\end{lstlisting}

The range is walked, the projection applied, and the results
gathered into an \code{ElementList}. Because \code{MapNode} produces
an \code{ElementList} (not a single \code{Element}), the parent
\code{v}/\code{h} splices its children into the parent's child list
--- the moral equivalent of a JSX fragment.

\subsection{The cost of \code{.build()}}

A pure compile-time tree (\code{t<"\dots">} and \code{v}/\code{h}
with no runtime children) does \code{.build()} in:

\begin{itemize}[leftmargin=*]
  \item One \code{std::string} construction per text node
    (small-string optimisation often makes this free).
  \item One \code{BoxElement} construction per box node, which holds
    a \code{std::vector<Element>} of children. The vector is
    reserved to the right size.
  \item Some \code{Style} field assignments.
\end{itemize}

No virtual dispatch, no template-method indirection. The whole tree
is built in a single recursive sweep that the compiler often inlines
completely for small trees. For a typical panel with twenty nodes,
\code{.build()} runs in tens of microseconds.

The \code{Element} tree is then dropped on the floor at the end of
the frame (after rasterisation into the canvas), and the next frame
builds a fresh one. This is the rebuild-every-frame discipline from
Chapter~\ref{ch:elm}.

\begin{recap}
\code{.build()} is the bridge from the compile-time tree to the
runtime \code{Element}. It is a one-shot recursive walk: every
node's \code{build()} produces an \code{Element}, children are
gathered into the parent's \code{Element}, and a runtime tree comes
out the bottom. Then it is the runtime side's problem.
\end{recap}

\section{Recipe gallery}
\label{sec:dsl:recipes}

A collection of patterns you will use constantly. Each is a small,
composable fragment; the goal is to make them feel familiar before
you need them.

\subsection{A titled panel with a border}

\begin{lstlisting}
constexpr auto title_panel = v(
    t<"Status"> | Bold | Underline,
    sep,
    t<"all systems nominal"> | Fg<80, 220, 120>
) | pad<1, 2> | border_<Round> | bcol<60, 65, 80>;
\end{lstlisting}

\code{sep} is a horizontal separator (a one-row box with a top
border); \code{pad<1, 2>} is one row of vertical padding, two
columns of horizontal.

\subsection{A two-column layout that grows to fit}

\begin{lstlisting}
auto split = h(
    sidebar  | w_<24>,
    content  | grow_<1>
) | grow_<1>;
\end{lstlisting}

The sidebar has fixed width 24; the content gets every remaining
column. The outer box grows to fill whatever vertical container is
holding it.

\subsection{A list of items from a runtime collection}

\begin{lstlisting}
auto list = v(
    t<"Files"> | Bold,
    map(files, [](const File& f) {
        return h(text(f.name) | w_<30>,
                 text(f.size) | Dim);
    })
);
\end{lstlisting}

The lambda is called once per file; each call produces an
\code{Element}. \code{map} gathers them into an \code{ElementList}
that \code{v} splices in.

\subsection{A conditional region}

Two forms, both common:

\begin{lstlisting}
// when(cond, then) --- hidden if cond is false
auto top = v(
    title,
    when(show_advanced, advanced_panel)
);

// when(cond, then, else) --- one or the other
auto status = when(loading,
    text("loading\dots") | Dim,
    text("ready")        | Fg<80, 220, 120>);
\end{lstlisting}

The one-argument form falls back to \code{BlankNode\{\}} (a
zero-cell empty text). The two-argument form picks one of two
branches; both branches must satisfy \code{Node}.

\subsection{A reusable style preset}

\begin{lstlisting}
constexpr auto heading = Bold | fg<0xFFFFFF>;
constexpr auto error   = Bold | fg<0xFF4444>;
constexpr auto subtle  = Dim  | Italic;

auto ui = v(
    t<"Welcome"> | heading,
    text(err)    | error,
    t<"v0.1">    | subtle
);
\end{lstlisting}

The presets are themselves \code{StyTag}s, computed at compile time.
Applying one is a single token; defining your design system this
way is the path to consistency.

\subsection{Inline width hints on text}

\begin{lstlisting}
text(very_long_string) | clip | w_<30>
\end{lstlisting}

\code{clip} switches the text node into truncate-with-ellipsis
mode. \code{w\_<30>} wraps the result in a fixed-width box, so the
text is forced to 30 columns regardless of available space.

\subsection{An animated spinner}

\begin{lstlisting}
dyn([&] {
    constexpr std::array frames = {"top", "upper-right", "lower-right", "lower-left"};
    return text(frames[(tick / 100) % frames.size()]);
}) | Fg<140, 220, 255>
\end{lstlisting}

The lambda picks a frame from the array based on the current tick
count. Combine with \code{Sub<Msg>::on\_animation\_frame(Tick\{\})}
from Chapter~\ref{ch:elm} for a 60\,Hz spinner.

\subsection{A right-aligned status bar}

\begin{lstlisting}
auto status_bar = h(
    text(left_text),
    space,                          // SpacerNode: grows to fill
    text(right_text) | Dim
);
\end{lstlisting}

\code{space} is the spacer node: a zero-content box with
\code{grow=1}. Between two non-growing siblings it stretches to
push them to opposite ends of the row.

\section{Why all of this, again?}
\label{sec:dsl:why-again}

At the end of a long chapter on templates, fold expressions, and
type-state machines, it is worth stepping back. What did all that
machinery buy us?

At build time:

\begin{itemize}[leftmargin=*]
  \item \textbf{No allocation for the tree shape.} The structure of
    your UI lives in types; only the leaf values (runtime strings,
    runtime children) allocate.
  \item \textbf{Errors are catch-able before the program runs.}
    Border colour without border, negative padding, wrong child
    type, hex overflow --- all rejected by the compiler.
  \item \textbf{The DSL composes with the host language.} Helper
    functions, lambdas, ranges --- all work without special
    integration.
\end{itemize}

At runtime:

\begin{itemize}[leftmargin=*]
  \item \textbf{\code{.build()} is one linear walk}, specialised by
    the compiler per distinct configuration. No virtual dispatch.
  \item \textbf{The runtime side is simple.} It receives a plain
    tree of \code{Element} values and lays it out; it does not
    care that the tree was generated by a compile-time DSL.
\end{itemize}

For build time. For correctness. For speed. Three things at once,
because we paid the cost in the DSL design.

\begin{idea}
The \maya{} DSL is what you get when you take ``the compiler should
catch this'' seriously and follow it all the way down. Strings,
styles, sizes, structure: in the types. Mistakes: caught at build.
Runtime: cheap and uniform. The trade is some up-front complexity in
the DSL itself; the payoff is every \maya{} program you ever write.
\end{idea}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Write a panel that shows your name in bold, an underlined
    tagline, and a divided list of three hobbies. Use only the
    compile-time pipes.
  \item Add a runtime counter (an \code{int} variable) with
    \code{dyn}. Increment it in a loop with
    \code{maya::live(...)} (covered in Chapter~\ref{ch:cmdsub}) and
    watch the panel refresh.
  \item Try \code{bcol<60, 65, 80>} on a box without a border. Read
    the compiler error. Find the line that says \code{constraints
    not satisfied} and the \code{Cfg.has\_border} that triggered it.
\end{enumerate}

\section{What is a compile-time program, really?}

The DSL is C++ code that runs while the compiler is running. The
result is a different C++ program, which is what your CPU
actually executes. Two languages, one source file. To use the DSL
well, it helps to see exactly where the line between the two
languages falls.

\subsection*{The two evaluation contexts}

C++ has, somewhat secretly, two evaluation contexts:

\begin{itemize}[leftmargin=*]
  \item \textbf{Compile-time} (\code{constexpr},
    \code{consteval}, template instantiation, non-type template
    parameters). Code that runs inside the compiler. Inputs and
    outputs are values known at the point the compiler reads
    them. No runtime, no heap (well, since C++20 there is a tiny
    one), no I/O.
  \item \textbf{Runtime}. Code that runs in the produced
    executable. Inputs come from arguments, files, the network,
    the user.
\end{itemize}

A \code{constexpr} function may run in \emph{either} context. A
\code{consteval} function may run only at compile time. A regular
function may run only at runtime (in general). The DSL uses all
three.

\subsection*{What survives into runtime}

Here is the rule that explains almost every DSL design choice:

\begin{quote}
A value known at compile time costs zero at runtime. A value not
known at compile time costs as much as it would have anyway.
\end{quote}

When you write \code{t<"hello"> | Bold}, the string \code{"hello"}
and the style \code{Bold} are both compile-time values. The pipe
is a \code{consteval} function that combines them into a
\code{TextNode} type whose template parameters encode both. When
the compiler emits machine code for that expression, there is no
string copy, no style construction, no allocation. Just a pointer
to a static byte sequence and a small struct on the stack.

When you write \code{text(my\_string) | Bold}, the string is a
runtime value. The pipe returns a \code{RuntimeTextNode} that
holds the string and the style as runtime fields. Now there is
an allocation (if the string is long), a copy, a struct.

The DSL gives you both spellings on purpose, because both costs
are appropriate sometimes. Spelling matters.

\begin{aside}
This is the same trade-off C++ string literals always had:
\code{"hello"} is a static byte sequence, \code{std::string\{"hello"\}}
is a heap-allocated copy. The DSL extends the trade to entire
UI fragments.
\end{aside}

\section{NTTPs, from zero}

A Non-Type Template Parameter (NTTP) is a value, not a type, that
appears in a template parameter list. You have seen them since
C++98:

\begin{lstlisting}
template <int N>
struct Array { int data[N]; };

Array<10> a;  // N is the value 10
\end{lstlisting}

For decades NTTPs were restricted to integral types, pointers,
and references. You could not pass a string literal, a struct,
or a floating-point number.

C++20 changed that. An NTTP may now be any \emph{structural
type}: a type with all-public data members, no user-defined
equality, trivially comparable. C++26 broadens the rules further.

\subsection*{Why this matters for a DSL}

With structural NTTPs, you can use a value as part of a type.
The key consequence: two instantiations with the same value
produce the same type. Two with different values produce
different types.

\begin{lstlisting}
template <Str s>  // Str is a structural type wrapping a literal
struct Tag {};

Tag<"hello">  a;
Tag<"hello">  b;  // same type as a
Tag<"world">  c;  // different type
\end{lstlisting}

This is the fundamental trick \maya{} uses to put strings,
styles, and box configurations into the type system. Once a
value is part of a type, the compiler can reason about it: it
can deduce, compare, dispatch, and overload.

\subsection*{The Str type}

\maya{}'s \code{Str} wraps a fixed-size character array so that a
string literal can become an NTTP:

\begin{lstlisting}
template <std::size_t N>
struct Str {
    char data[N];
    constexpr Str(const char (&s)[N]) {
        for (std::size_t i = 0; i < N; ++i) data[i] = s[i];
    }
    constexpr std::string_view view() const {
        return {data, N - 1};  // strip trailing \0
    }
};
\end{lstlisting}

A literal \code{"hello"} has type \code{const char[6]}. The
constructor copies its bytes into the \code{data} array of a
\code{Str<6>} value. Two \code{Str<6>} instances with the same
bytes are equal under member-wise comparison, so they are the
same NTTP value, so they instantiate the same type.

This is what makes \code{t<"hello">} work: \code{t} is a template
that takes a \code{Str} NTTP, and the literal is implicitly
constructed.

\begin{funfact}
The \code{Str} pattern is sometimes called \emph{cstring NTTP}
or \emph{template string literal}. It first showed up in C++20
proposals and is now widely used in libraries like fmt (for
compile-time format-string checking).
\end{funfact}

\subsection*{Compile-time style: CTStyle}

The same trick gives us compile-time styles. \code{CTStyle} is a
structural type holding flags and colours:

\begin{lstlisting}
struct CTStyle {
    bool bold      = false;
    bool dim       = false;
    bool italic    = false;
    bool underline = false;
    // ... fg, bg, ...
};
\end{lstlisting}

A style tag is a template wrapping a \code{CTStyle} NTTP:

\begin{lstlisting}
template <CTStyle S>
struct StyTag {};

constexpr StyTag<CTStyle{.bold = true}>       Bold;
constexpr StyTag<CTStyle{.italic = true}>     Italic;
constexpr StyTag<CTStyle{.underline = true}>  Underline;
\end{lstlisting}

Now \code{Bold} is a value of a specific type, distinct from
\code{Italic}, and combining them by pipe is overload resolution
on types --- not a runtime branch.

\subsection*{Composing styles: pipe between StyTags}

What does \code{Bold | Italic} mean? It is operator-pipe applied
to two style tags:

\begin{lstlisting}
template <CTStyle A, CTStyle B>
constexpr auto operator|(StyTag<A>, StyTag<B>) {
    constexpr CTStyle merged = merge(A, B);
    return StyTag<merged>{};
}
\end{lstlisting}

\code{merge} is a \code{constexpr} function that walks the two
\code{CTStyle}s field by field and produces a third. The whole
computation happens in the compiler; the resulting program just
has a single \code{StyTag} value of the merged type.

\subsection*{Truecolour: Fg and Bg as templates}

Foreground and background colours arrive the same way:

\begin{lstlisting}
template <uint8_t R, uint8_t G, uint8_t B>
constexpr auto Fg = StyTag<CTStyle{.fg = {R, G, B}}>{};
\end{lstlisting}

So \code{Fg<255, 100, 50>} is a style tag whose type encodes
an exact RGB triple. Combining it with \code{Bold} gives a tag
that encodes both --- still entirely in the type system.

The convenience macro \code{fg<"\#FF6432">} parses a hex literal
at compile time. The parser is a \code{constexpr} function over
a \code{Str}; the output is the same \code{Fg<R, G, B>}.

\begin{warning}
MSVC used to (and may still, depending on version) refuse to
deduce \code{Fg} as a variable template across some translation
units. \maya{}'s workaround is the \code{detail\_rgb} namespace
--- an explicit factory function that MSVC handles. If you see
surprising errors only on Windows, this is usually the cause.
\end{warning}

\section{Every pipe overload, by left-hand-side type}

The pipe \code{|} is overloaded many times. Each overload is
selected by the type of the left operand. Below is the full
table of overloads, organised by LHS.

\subsection*{LHS: TextNode (compile-time text)}

\begin{itemize}[leftmargin=*]
  \item \code{TextNode<S> | StyTag<St>} $\to$ \code{TextNode<S, St>}.
    Attaches style to the text. Compile-time only.
  \item \code{TextNode<S> | pad\_t<P>} $\to$ \code{BoxNode<\ldots>}.
    Wraps the text in a padded box. The padding is compile-time.
  \item \code{TextNode<S> | width\_t<W>} $\to$
    \code{BoxNode<\ldots>}. Clips/extends the text to a fixed
    width.
\end{itemize}

\subsection*{LHS: RuntimeTextNode}

\begin{itemize}[leftmargin=*]
  \item \code{RuntimeTextNode \& | StyTag<St>} $\to$
    \code{RuntimeTextNode} with style copied in. Runtime fields
    get a runtime style attached. The compile-time style is
    flattened into the runtime struct.
  \item \code{RuntimeTextNode \& | pad\_t<P>} $\to$ runtime box.
\end{itemize}

\subsection*{LHS: BoxNode}

A box node can take more decoration than a leaf text node, so it
has many pipe overloads:

\begin{itemize}[leftmargin=*]
  \item \code{BoxNode<Cfg> | border\_t<B>}. Adds a border. Only
    valid if \code{Cfg.has\_border == false} (cannot add a border
    twice). Compile error otherwise.
  \item \code{BoxNode<Cfg> | bcol<R, G, B>}. Sets border colour.
    Only valid if \code{Cfg.has\_border == true}. Compile error
    otherwise.
  \item \code{BoxNode<Cfg> | pad\_t<P>}. Adds padding inside the
    box.
  \item \code{BoxNode<Cfg> | width\_t<W>}. Sets a fixed width.
  \item \code{BoxNode<Cfg> | height\_t<H>}. Sets a fixed height.
  \item \code{BoxNode<Cfg> | StyTag<St>}. Applies style to the
    box's content.
  \item \code{BoxNode<Cfg> | flex\_t<G>}. Sets flex-grow factor.
\end{itemize}

Each overload is gated by a \code{requires} clause that checks
the \code{Cfg} for compatibility. The clauses are what produce
the friendly compile errors.

\subsection*{LHS: MapNode, WhenNode, VisibleNode}

These wrappers carry through the pipe overloads from their
inner type. \code{when(cond, child) | Bold} is equivalent to
\code{when(cond, child | Bold)}. The wrapping is transparent.

\begin{recap}
The pipe is not one operator; it is dozens. Each overload is
specific to a (LHS, RHS) pair. The compiler picks the right one
by type. There is no virtual dispatch, no runtime branching ---
the whole pipeline is one type, fully resolved by the time the
backend sees it.
\end{recap}

\section{The type-state machine}

The DSL is a \emph{type-state machine}: each node type encodes
what operations are currently legal on it. As you compose, the
type changes; the set of legal next operations changes with it.

Think of \code{BoxNode<Cfg>} as a state. The \code{Cfg} contains
flags like \code{has\_border}, \code{has\_width},
\code{has\_border\_colour}. Each pipe overload requires certain
flags to be set or unset, and (if it succeeds) flips one or more
flags in the output type.

A transition diagram:

\begin{lstlisting}
start: BoxNode<{has_border=0, has_bcol=0, has_pad=0}>
  | border<Rounded>   --> BoxNode<{has_border=1, has_bcol=0, has_pad=0}>
  | bcol<60,65,80>    --> compile error (has_border==0)

after border:
  BoxNode<{has_border=1, has_bcol=0, has_pad=0}>
  | bcol<60,65,80>    --> BoxNode<{has_border=1, has_bcol=1, has_pad=0}>
  | border<Double>    --> compile error (has_border==1)
  | pad<1>            --> BoxNode<{has_border=1, has_bcol=1, has_pad=1}>
\end{lstlisting}

Illegal moves do not become runtime errors --- they fail to
compile, with a message naming the unsatisfied constraint. This
is what we mean when we say ``the DSL pushes errors into the
build''.

\begin{idea}
Type-state has been used in Rust's typestate proposals, in
Phantom-typed APIs in Haskell, and in fluent builders in Java
(``you can only call \code{build()} after setting all required
fields''). \maya{}'s version uses C++26 NTTPs to make the state a
first-class structural value instead of a phantom tag.
\end{idea}

\section{Decoding compile errors, expanded}

When a constraint fails, the compiler produces a message. Modern
compilers (GCC 13+, Clang 17+) print the unsatisfied \code{requires}
clause inline. Here are five real errors and how to read them.

\subsection*{Error 1: border colour without border}

Source:

\begin{lstlisting}
auto bad = box(text("hi")) | bcol<60, 65, 80>;
\end{lstlisting}

GCC error excerpt:

\begin{lstlisting}
error: no match for 'operator|' (operand types are 'BoxNode<...>'
                                  and 'bcol_t<60, 65, 80>')
note: candidate: 'requires (Cfg.has_border) ...'
note:   constraint not satisfied: 'Cfg.has_border == true'
\end{lstlisting}

Read: ``the candidate \code{operator|} for box+border-colour
exists, but its \code{requires} clause needs
\code{Cfg.has\_border == true}, and your box has
\code{has\_border == false}.'' Fix: pipe \code{border<...>} in
first.

\subsection*{Error 2: negative padding}

Source:

\begin{lstlisting}
auto bad = text("hi") | pad<-1>;
\end{lstlisting}

Error excerpt:

\begin{lstlisting}
error: template argument deduction/substitution failed:
note:   value '-1' does not satisfy 'PadValue' (must be >= 0)
\end{lstlisting}

Read: ``the \code{pad} template expects a non-negative value;
$-1$ fails the concept check.'' Fix: use a non-negative literal,
or compute the padding at runtime via \code{dyn\_pad(\ldots)}.

\subsection*{Error 3: wrong child type}

Source:

\begin{lstlisting}
auto bad = h(42, text("hi"));
\end{lstlisting}

Error excerpt:

\begin{lstlisting}
error: no matching function for call to 'h(int, TextNode<...>)'
note:   constraint not satisfied: 'ElementLike<int>'
\end{lstlisting}

Read: ``\code{h} (horizontal layout) requires all children to
satisfy the \code{ElementLike} concept; \code{int} does not.''
Fix: wrap with \code{text(std::to\_string(42))} if you wanted to
render the number.

\subsection*{Error 4: a Str that is not constexpr}

Source:

\begin{lstlisting}
std::string s = compute_label();
auto bad = t<s>;
\end{lstlisting}

Error excerpt:

\begin{lstlisting}
error: 's' is not a constant expression
note:   's' was not declared 'constexpr'
\end{lstlisting}

Read: ``the template argument must be a constant expression;
\code{s} is not.'' Fix: use \code{text(s)} instead --- the runtime
version.

\subsection*{Error 5: pipe order ambiguity}

Source:

\begin{lstlisting}
auto bad = text("hi") | width<10> | Bold | width<20>;
\end{lstlisting}

Error excerpt:

\begin{lstlisting}
error: 'width' has already been set on this BoxNode
note:   each modifier may be applied at most once
\end{lstlisting}

Read: ``the type-state forbids setting \code{width} twice.''
Fix: keep one \code{width}. If you really want to override,
create a new node and pipe again.

\subsection*{A reading strategy}

When the DSL produces an error:

\begin{enumerate}[leftmargin=*]
  \item Find the line that says \code{constraint not satisfied}
    or \code{candidate ignored}. This names the failing rule.
  \item Read the rule. It is usually a short \code{requires}
    clause that mentions one or two flags or concepts.
  \item Map the rule back to your source. The compiler usually
    points at the operator, not the offending pipe; the offending
    pipe is the previous one.
  \item Fix the order, or fix the value, or move to a runtime
    variant.
\end{enumerate}

\begin{aside}
If the error trace is overwhelming, set \code{-fconcepts-diagnostics-depth=2}
on GCC. The deep depth dumps every level of template substitution;
depth 2 keeps the message focused on the user-visible constraint.
\end{aside}

\section{What .build() does, per node kind}

Every node has a \code{build()} method that converts the
compile-time tree into a runtime \code{Element}. The work it does
depends on the node kind. Tracing the cost by kind reveals where
the DSL's runtime overhead actually lives.

\subsection*{TextNode<S, St>::build()}

The text is a literal, the style is a value. \code{build} produces
an \code{Element\{Text\{view\_of\_S, runtime\_style\_of\_St\}\}}.
No allocation; the view points into the static string and the
style is a small POD.

Cost: one struct construction. Microseconds for a screen of text.

\subsection*{RuntimeTextNode::build()}

The text is an owned \code{std::string}, the style is a value.
\code{build} moves the string into the \code{Text} variant and
attaches the style. If the string was small (SSO), no allocation;
otherwise the original heap buffer is transferred.

Cost: a move (cheap) or, if you handed in a temporary,
zero moves.

\subsection*{BoxNode<Cfg>::build()}

The most complex case. \code{build} walks the children, calling
\code{build()} on each, then assembles a \code{Box} struct with
the assembled children, the layout style derived from \code{Cfg},
and any decoration (border, padding, colour).

Cost: $O(n)$ in the number of descendant nodes, plus one
allocation for the children vector. For a typical UI tree of a
few hundred nodes, this is on the order of microseconds.

\subsection*{MapNode<F>::build()}

Map is a wrapper that lifts child messages through a function.
\code{build} calls the child's \code{build}, then attaches the
function so that the runtime event system can apply it. The
function itself is a \code{std::function} (so there is one
allocation for the closure if it captures).

\subsection*{WhenNode::build()}

A conditional. If the condition is \code{true}, build the child;
otherwise return an empty \code{Element}. The runtime cost is
zero when the condition is false; when true, it is the cost of
the child.

\subsection*{VisibleNode::build()}

Like \code{WhenNode} but the child is always built and only its
visibility flag is set. Use when you want the layout to reserve
space even when the element is hidden.

\begin{recap}
The runtime cost of the DSL is one tree walk and a handful of
allocations per frame. The compile-time machinery is invisible at
runtime. If you measure your frame and it is microseconds, the
DSL is doing its job.
\end{recap}

\section{Recipe gallery}

A tour of small, useful patterns. Each is one screen of code and
shows a different facet of the DSL.

\subsection*{Titled panel}

\begin{lstlisting}
auto panel = box(
    v(text("Title") | Bold | Underline,
      t<"---------"> | Dim,
      text("body line 1"),
      text("body line 2")))
    | border<Rounded> | pad<1>;
\end{lstlisting}

A border, padding, and a divided header. All compile-time; the
runtime sees a fully assembled tree.

\subsection*{Two-column layout}

\begin{lstlisting}
auto two_col = h(
    box(text(left_text))  | flex<1>,
    t<" | "> | Dim,
    box(text(right_text)) | flex<1>);
\end{lstlisting}

Two flex children share the row equally. The divider takes its
intrinsic width.

\subsection*{Status bar}

\begin{lstlisting}
auto bar = h(
    text(mode_label)  | Bold | Bg<0,80,150>,
    text(" " + file)  | flex<1>,
    text(line_col)    | Dim)
    | height<1>;
\end{lstlisting}

A fixed-height row with three regions: a coloured mode tag, a
flexible file name, and a dim line/column indicator.

\subsection*{Spinner}

\begin{lstlisting}
constexpr std::array frames = {
    "upper-left", "upper-right", "lower-right", "lower-left"};
auto spinner = text(frames[tick % 4]) | Bold;
\end{lstlisting}

A tiny animation. The frame list is static; the indexing is
runtime; the styling is compile-time.

\subsection*{Conditional element}

\begin{lstlisting}
auto maybe_help = when(show_help,
    box(text("press q to quit")) | border<Single> | pad<1>);
\end{lstlisting}

If \code{show\_help} is false, the element costs nothing to
render. If true, the panel appears.

\subsection*{List with separators}

\begin{lstlisting}
auto list = v(
    items                  // a std::vector<Element>
    | std::views::join_with(t<" --- "> | Dim));
\end{lstlisting}

Using C++23 ranges, intersperse a divider between elements. The
divider is compile-time, the items are runtime.

\subsection*{Style preset}

\begin{lstlisting}
constexpr auto Heading = Bold | Underline | Fg<200, 220, 255>;
auto h1 = text("Chapter 1") | Heading;
auto h2 = text("Chapter 2") | Heading;
\end{lstlisting}

A reusable style. \code{Heading} is itself a \code{StyTag} whose
type encodes all three properties. The pipe combines them
lexically; the runtime sees one merged style.

\subsection*{Width-clipped text}

\begin{lstlisting}
auto clipped = text(very_long_string) | width<40>;
\end{lstlisting}

Forces the text to occupy exactly 40 columns, truncating with an
ellipsis if longer. Good for table cells and status lines.

\section{When the DSL is not enough}

Not every UI shape fits the compile-time pipeline:

\begin{itemize}[leftmargin=*]
  \item \textbf{Dynamic tree structure.} If you do not know the
    set of children until runtime (e.g. a file tree), you build
    a \code{std::vector<Element>} at runtime and pass it to
    \code{v(\ldots)}. The static pipeline still wraps the dynamic
    leaves.
  \item \textbf{User-themable styles.} If the user picks colours
    in a settings menu, you cannot encode the colour in the type.
    Use the runtime \code{Style} struct and apply it with
    \code{styled(\ldots)}.
  \item \textbf{Loaded layouts.} If the layout comes from a
    config file, the structure cannot be encoded statically.
    The DSL is still useful for sub-trees; the top-level becomes
    a hand-built \code{Element}.
\end{itemize}

The DSL is one of several tools. Use it where types can carry
the weight; fall back to plain \code{Element} construction where
they cannot.

\begin{warning}
There is a temptation to encode \emph{everything} at compile time.
Resist it for dynamic data. The runtime variants exist for a
reason --- they will save you from a wall of
\code{constexpr}-related compile errors when the requirement
changes.
\end{warning}
