\chapter{Flexbox in integers}
\label{ch:layout}

Given an element tree, \maya{} must decide where every text run, every
box, every border goes --- in cells, not pixels. The algorithm it uses
is a tailored re-implementation of CSS Flexbox. This chapter explains
flexbox from scratch (no web background assumed) and then walks
\maya{}'s implementation.

\begin{aside}
Layout is the unglamorous middle of every UI library. Nobody talks
about it; the algorithm runs hundreds of times a second, makes
thousands of small decisions, and disappears the moment it gets the
rectangle right. Bad layout is instantly noticeable; good layout is
invisible. This chapter is a guided tour of \maya{}'s ten kilobytes
of layout code, because once you understand it you can stop guessing
at why a panel is one cell off and start reading the cause out of
the configuration.
\end{aside}

\section{The problem layout solves}

You have a rectangular area --- the terminal --- and a tree of
elements each with constraints (\emph{this one wants to be 20 cells
wide; this row should pack its children with a 1-cell gap; this panel
must grow to fill leftover space}). Layout is the function that
assigns concrete \code{(x, y, w, h)} rectangles to each node such that
all the constraints are satisfied.

There are many possible algorithms (table layout, constraint solving,
grid layout). Flexbox is the most appropriate for terminal UIs:
predictable, single-pass, no surprises, no solver.

\subsection{Why not the obvious algorithms?}

A brief tour of the alternatives \maya{} did not choose:

\paragraph{Table layout (HTML \code{<table>}).} A rectangular grid
of rows and columns. Each cell's size is determined by the largest
content in its row and column. Beautiful for tabular data; awkward
for anything else --- two unrelated cells affecting each other's
size is a constant surprise. Used in early web design; abandoned
as a general-purpose layout in the mid-2000s.

\paragraph{Constraint solving (Cassowary, Auto Layout).} Each
position is a variable; each layout rule is a linear constraint.
A simplex-like solver finds an assignment. Powerful (it can express
``these two are always equal'', ``this is between 10 and 30
cells''), but expensive and prone to producing surprising answers
when constraints conflict. Apple's Auto Layout works this way.

\paragraph{Grid (CSS Grid).} A two-dimensional placement model:
you declare a grid of named rows and columns, then place children
in specific cells. Excellent for page-level layout; overkill for
the nested-row-and-column shape most TUIs take.

\paragraph{Flow / line-box (HTML inline text).} Children are
packed left-to-right, top-to-bottom, like words in a paragraph.
Good for prose; bad for everything else.

Flexbox is the simplest model that handles the common cases ---
rows of widgets, columns of panels, a sidebar plus content, a
status bar pinned to the bottom --- without ceremony. It is also
linear-time and predictable, both crucial for a sixty-frames-per-second
renderer.

\begin{funfact}
Flexbox went through several names and standards before it stuck.
The original CSS3 ``flexible box'' module was proposed in 2009;
the syntax changed twice (the \code{display: box} from 2009, then
\code{display: flexbox} in 2011, then \code{display: flex} in 2012)
before browsers settled. Mosaic's table-based layouts and
Netscape's \code{<center>} tag ran the web for a decade because
nothing nicer was standardised. The model \maya{} uses --- main
axis, grow, shrink, basis, gap --- is the 2012 design, late enough
that the rough edges had been filed off, simple enough to fit in
800 lines of \cpp{}.
\end{funfact}

\section{Flexbox in one paragraph}

Flexbox arranges children along a \textbf{main axis} (row or column).
Children may \emph{grow} to fill leftover space and \emph{shrink} when
oversized; they may \emph{wrap} onto new lines when they don't fit;
they may be \emph{aligned} along the cross axis and \emph{justified}
along the main axis. Each container picks an axis (row or column);
each child carries a few numbers (basis, grow, shrink, gap, alignment
override).

\section{Vocabulary}

A worked example will be clearest. Take a horizontal row of three
children inside a 30-cell-wide box:

\begin{lstlisting}
h(
    panel_a,   // wants 10 cells
    panel_b,   // grow=1
    panel_c    // wants 8 cells
) | gap_<1>
\end{lstlisting}

\maya{} reasons about it like this:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Sum the basis sizes.} A=10, B=0 (no fixed size,
    grows), C=8. Plus 2 gap cells. Total committed: 20.
  \item \textbf{Leftover space.} 30 - 20 = 10 cells.
  \item \textbf{Distribute by grow.} Only B has \code{grow > 0}, so
    B receives all 10 cells. New size: B = 10.
  \item \textbf{Place.} A at column 0, gap, B at column 11, gap, C
    at column 22.
\end{enumerate}

The cross axis (vertical, in this example) is handled by
\code{Align}: children stretch to the parent height by default, but
they can be \code{Start}, \code{Center}, \code{End}, or
\code{Stretch}.

\section{The flex properties}

A node in \maya{}'s layout engine carries the following style:

\begin{lstlisting}
struct FlexStyle {
    FlexDirection direction = FlexDirection::Column;
    FlexWrap      wrap      = FlexWrap::NoWrap;

    Align   align_items   = Align::Auto;
    Align   align_self    = Align::Auto;
    Justify justify       = Justify::Auto;

    int     gap           = 0;

    int     pad_top    = 0, pad_right = 0,
            pad_bottom = 0, pad_left  = 0;
    int     mar_top    = 0, mar_right = 0,
            mar_bottom = 0, mar_left  = 0;

    int     width       = -1;   // -1 = auto
    int     height      = -1;
    int     min_width   = 0, max_width  = INT_MAX;
    int     min_height  = 0, max_height = INT_MAX;

    int     basis       = -1;   // preferred main-axis size, -1 = auto
    float   grow        = 0.0f;
    float   shrink      = 1.0f;

    Display display     = Display::Flex;
    Overflow overflow   = Overflow::Visible;
    bool    has_border  = false;
};
\end{lstlisting}

These mirror CSS flexbox almost exactly. The differences:

\begin{itemize}[leftmargin=*]
  \item Every quantity is an \emph{integer} number of cells. No
    \code{em}, no \code{px}, no fractional pixels. Sub-cell positions
    don't exist in a terminal.
  \item \code{display: Flex} and \code{display: None} are the only
    options. There is no block, no inline.
  \item Percentages are limited; most sizes are explicit integers.
\end{itemize}

\section{The data structure}

A layout tree is a flat \code{std::vector<LayoutNode>} indexed by
\code{int}. Each node has a list of child indices.

\begin{lstlisting}
struct LayoutNode {
    FlexStyle           style;
    std::vector<int>    children;     // indices into the same vector
    layout::Rect        computed;     // output: x, y, w, h
};
\end{lstlisting}

Why a flat vector and not a pointer tree? Three reasons:

\begin{itemize}[leftmargin=*]
  \item \textbf{Cache locality.} Walking siblings is a stride of
    \code{sizeof(LayoutNode)}; the prefetcher loves it.
  \item \textbf{No allocation in the hot path.} A
    \code{vector<int>} for children is cheaper than a \code{vector
    <unique\_ptr<LayoutNode>>}.
  \item \textbf{Easier reasoning.} Indices cannot dangle; an index
    is either in range or out of range. Pointer trees create
    aliasing puzzles.
\end{itemize}

\begin{idea}
``Make it an index, not a pointer'' is a recurring optimisation in
\maya{}. The same idea drives the style pool (a 16-bit ID into a
table) and the hyperlink interning (a 32-bit ID).
\end{idea}

\section{The algorithm, step by step}

\code{layout::compute(nodes, root, available\_width)} runs in three
phases per box:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Measure.} Walk children recursively and compute their
    intrinsic main-axis size (the basis if specified, else the
    content's measured size).
  \item \textbf{Distribute.} Add up basis sizes plus gaps. Compare
    to the parent's main-axis size. If there is leftover space and
    any child has \code{grow > 0}, distribute the leftover
    proportionally; if there is overflow and any child has
    \code{shrink > 0}, shrink proportionally.
  \item \textbf{Place.} Compute final \code{(x, y, w, h)} for each
    child given the resolved main-axis sizes and the cross-axis
    alignment.
\end{enumerate}

Wrapping (when \code{wrap = Wrap}) re-enters the algorithm per line.
Justify and align use simple integer division: \code{SpaceBetween}
puts the leftover between children, \code{SpaceAround} puts it on
both sides, \code{Center} halves it.

\subsection{Padding, border, margin}

Three rectangles concentric:

\begin{center}
\small
\begin{tabular}{ll}
\textbf{margin} & cells \emph{outside} the box (sibling spacing) \\
\textbf{border} & one cell thick if present (drawn glyphs) \\
\textbf{padding} & cells \emph{inside} the border (content inset) \\
\textbf{content} & where children are laid out \\
\end{tabular}
\end{center}

The available width for children is \code{w - left\_pad - right\_pad -
border\_l - border\_r}. The available height is the same on the
vertical. \maya{}'s renderer paints the border in the one-cell ring
around the padded content.

\section{Worked example: a status panel}

\begin{lstlisting}
auto panel = v(
    t<"Status"> | Bold,
    sep,
    h(t<"CPU:"> | Dim, space, t<"42%"> | Bold) | gap_<1>,
    h(t<"Mem:"> | Dim, space, t<"8.2G"> | Bold) | gap_<1>
) | border_<Round> | pad<1, 2> | gap_<1>;
\end{lstlisting}

Suppose the parent box gives us 24 cells wide. The layout proceeds:

\begin{enumerate}[leftmargin=*]
  \item Outer box: border (1 cell each side) + padding (2 on each
    side) leaves 24 - 2 - 4 = \textbf{18 cells} of inner width.
  \item Inner column has four children: title (1 row), separator (1
    row), CPU line (1 row), Mem line (1 row), with gap 1 between
    them. Total height 4 + 3 = 7 rows of content. Plus the border
    pads (1 + 1) and padding (1 + 1) makes the outer box 11 rows.
  \item Inside each \code{h(...)} line, the children are: a static
    label, a \code{space} (grow=1), and a value text. Total fixed
    width: maybe 4 + 3 + gap 1 = 8 cells. Leftover 10 goes entirely
    to the spacer. The value text is pushed to column 8 + 10 = 18,
    i.e. the right edge.
\end{enumerate}

The renderer then walks the resulting rectangles and writes glyphs.

\section{When layout disagrees with you}

Two situations confuse newcomers:

\begin{itemize}[leftmargin=*]
  \item \textbf{Nothing happens when I set \code{grow}.} The parent
    must have a fixed (or grow-controlled) size for the leftover
    space to be meaningful. A \code{grow=1} child of a parent that
    sizes itself to its content has nothing to grow into.
  \item \textbf{Text wraps inside a column unexpectedly.} A
    \code{TextElement}'s width is its content. If you put a long
    string inside a column, it tries to take its natural width; the
    parent box constrains it to its own width, which causes wrap.
    Set \code{TextWrap::Truncate} or \code{Ellipsize} if that is
    wrong.
\end{itemize}

\section{Where the code lives}

\file{maya/include/maya/layout/yoga.hpp} and
\file{maya/src/layout/yoga.cpp}. Despite the name, this is not a
binding to Facebook's Yoga library --- it is an independent
implementation that follows the same flexbox spec. ``Yoga'' here is a
nickname for the algorithm. The implementation is around 800 lines
and has no external dependencies.

\section{Justify and Align, every value}
\label{sec:layout:justify-align}

The two enums \code{Justify} and \code{Align} control how leftover
space is distributed and how children line up across the cross axis.
Both share a value list: \code{Start}, \code{End}, \code{Center},
\code{Stretch}, \code{SpaceBetween}, \code{SpaceAround},
\code{SpaceEvenly}, \code{Auto}. Worth walking each.

Imagine a row 30 cells wide, holding three children each 4 cells
wide. Leftover space: $30 - 12 = 18$ cells.

\paragraph{\code{Start} (default).} Pack children to the start.
No space inserted between them; all 18 leftover cells go to the
end.

\begin{terminalbox}
AAAA BBBB CCCC..................
\end{terminalbox}

\paragraph{\code{End}.} Pack children to the end. All leftover
space at the start.

\begin{terminalbox}
..................AAAA BBBB CCCC
\end{terminalbox}

\paragraph{\code{Center}.} Half the leftover before, half after.
With odd leftover (rare in cells but possible), the extra cell
goes to the start.

\begin{terminalbox}
.........AAAA BBBB CCCC.........
\end{terminalbox}

\paragraph{\code{SpaceBetween}.} Leftover divided into
$n - 1$ equal gaps between children. First child at start, last at
end. With 18 cells divided into 2 gaps: 9 each.

\begin{terminalbox}
AAAA.........BBBB.........CCCC
\end{terminalbox}

\paragraph{\code{SpaceAround}.} Leftover divided into $n$ equal
segments, half a segment on each end, full segments between.
With 18 cells over 3 children: 6 per segment, 3 on each end.

\begin{terminalbox}
...AAAA......BBBB......CCCC...
\end{terminalbox}

\paragraph{\code{SpaceEvenly}.} Leftover divided into $n + 1$
equal gaps. Same gap on each end as between children. With 18
cells over 4 gaps: 4 or 5 each (integer division leaves a remainder).

\begin{terminalbox}
....AAAA....BBBB....CCCC...
\end{terminalbox}

\paragraph{\code{Stretch}.} Cross-axis only. Children are sized to
the parent's cross-axis size. The default for cross-axis alignment
on boxes (so children fill the height of a row by default).

\paragraph{\code{Auto}.} The default. For \code{align\_items}
resolves to \code{Stretch}; for \code{align\_self} means ``inherit
from parent's \code{align\_items}''.

\subsection{Integer division and the remainder}

A wrinkle of cell-based layout: leftover space rarely divides
evenly. With \code{SpaceBetween} of three children in 30 cells
(leftover 18, two gaps), each gap is 9. With four children in 30
cells (leftover 14, three gaps), each gap is 4 and the integer
division leaves 2 cells unaccounted for.

\maya{}'s strategy: the first $k$ gaps get one extra cell, where
$k$ is the remainder. Children later in the list see a slightly
tighter gap. The alternative would be to round each gap and
accept that the total drifts by a cell or two, but accumulating
rounding errors quickly produces visible misalignment.

\begin{idea}
In integer cell-based layout, leftover space rarely divides
evenly. The convention is to give the remainder to the early gaps,
so the off-by-one shows up at one edge rather than accumulating.
This is a recurring theme in pixel-free graphics; the same
problem appears in audio sample rate conversion and frame-rate
remapping.
\end{idea}

\section{Grow and shrink, mechanically}
\label{sec:layout:grow-shrink}

The \code{grow} and \code{shrink} floats are the most subtle part
of flexbox. Worth walking through their full mechanics, with
worked examples.

\subsection{The setup}

Given children with basis sizes $b_1, b_2, \dots, b_n$ and an
available main-axis size $W$:

\begin{enumerate}[leftmargin=*]
  \item Sum the basis sizes and gaps: $B = \sum b_i + \text{gaps}$.
  \item If $B = W$, you are done. Place children at their basis
    sizes.
  \item If $B < W$: there is leftover space, and \textbf{grow}
    distributes it. If no child has \code{grow > 0}, leftover stays
    leftover (handled by \code{justify}).
  \item If $B > W$: children overflow, and \textbf{shrink}
    distributes the negative. If no child has \code{shrink > 0},
    the overflow stays (visible as content escaping the box, or
    clipped depending on \code{overflow}).
\end{enumerate}

\subsection{Distributing leftover by \code{grow}}

Let $G = \sum g_i$ be the total grow factor of all children. If
$G > 0$ and $L = W - B > 0$, each child $i$ with grow $g_i$ receives
an additional $\lfloor L \cdot g_i / G \rfloor$ cells.

A worked example. Three children, basis sizes 10, 5, 5 (total 20),
grow factors 1, 0, 2 (total 3), available 32 cells. Leftover
$L = 12$. Distribute:

\begin{itemize}[leftmargin=*]
  \item Child 1: $\lfloor 12 \cdot 1 / 3 \rfloor = 4$ extra cells.
    Final size 14.
  \item Child 2: grow 0, no extra. Final size 5.
  \item Child 3: $\lfloor 12 \cdot 2 / 3 \rfloor = 8$ extra cells.
    Final size 13.
\end{itemize}

Total: $14 + 5 + 13 = 32$, exactly $W$. The remainders from
division sum to zero in this case; when they don't, the first child
in the list absorbs the difference (so the total comes out exact).

\subsection{Distributing overflow by \code{shrink}}

The symmetric case. If $B > W$ and total shrink $S > 0$, each child
loses $\lfloor (B - W) \cdot s_i / S \rfloor$ cells. The defaults
are important: \code{shrink = 1.0f} means every child shrinks
equally; \code{shrink = 0.0f} means the child refuses to shrink and
overflow goes elsewhere.

A child whose basis is already at its \code{min\_width} cannot
shrink further; the algorithm marks it ineligible and redistributes
over the remaining children. This is one of two places \maya{}'s
flexbox implementation does a second pass.

\subsection{Common patterns}

The spinner of common grow/shrink configurations:

\begin{center}
\small
\begin{tabular}{ll}
\textbf{Effect} & \textbf{Setting} \\
\hline
Fixed size, never resizes        & \code{grow=0, shrink=0, basis=N} \\
Fills leftover space              & \code{grow=1, shrink=1} (defaults) \\
Fills, never overflows            & \code{grow=1, shrink=1, min\_width=0} \\
Fills proportionally (2:1 split)  & two siblings with \code{grow=2} and \code{grow=1} \\
Never shrinks below content       & \code{shrink=0} \\
Doesn't grow, will shrink         & \code{grow=0, shrink=1} \\
\end{tabular}
\end{center}

\code{space} in the DSL is a node with \code{grow=1, basis=0,
shrink=0}: it eagerly absorbs every leftover cell and refuses to be
shrunk if there isn't enough space.

\section{Why integer cells, not floats}
\label{sec:layout:integer-math}

Web flexbox uses floating-point pixels because the visual unit ---
a pixel --- can effectively be subdivided in CSS (sub-pixel
antialiasing, fractional positions, transforms). The terminal has
no such unit: a cell is a cell, you cannot draw half a character.
Layout has to terminate on whole cells.

Using \code{int} for every quantity has consequences worth
being explicit about:

\paragraph{No rounding surprises during composition.} Two layouts
computed independently and then combined have exactly the same
geometry as if computed together. With floats, rounding error
compounds: a panel that is 12.499... cells wide at one level
becomes 12 at the next, then 11.998... in the parent, and so on.
Integer math avoids the question.

\paragraph{Diff is bitwise comparable.} Frame-to-frame layout
diffs (Chapter~\ref{ch:simd}) can compare integer rectangles
with a single \code{==}; with floats the question of ``did this
position change'' becomes ``did this position change by more than
epsilon'', and the epsilon has to be chosen.

\paragraph{Constraints are exact.} ``This box is at least 10 cells
wide and at most 30'' is bit-exact; the solver never needs to
worry about a width that is mathematically 10.0001 cells.

\paragraph{The cost: integer division does not preserve totals.}
The technique from the previous section --- give remainders to the
earliest gaps --- is what we pay for using integers. It is much
less work than the floating-point sorrows we avoid.

\section{The three phases, expanded}
\label{sec:layout:phases}

The summary said ``measure, distribute, place''. The full version
for a single box:

\paragraph{Phase 1: measure.}
For each child, recursively compute its \emph{intrinsic} main-axis
size --- the size it wants to be if given no extra space.
\code{TextElement}'s intrinsic main-axis size is the width of its
content (for a row container) or the height after wrapping (for a
column). \code{BoxElement}'s intrinsic is its own measured size,
computed by recursing into its children.

The result is each child's \textbf{basis} --- either the explicit
\code{basis} field if set, or the measured intrinsic.

\paragraph{Phase 2: distribute.}
Compare $B = \sum b_i + \text{gaps}$ to $W$, the available
main-axis size. Distribute leftover or overflow per
Section~\ref{sec:layout:grow-shrink}. After this phase every
child has a final main-axis size.

\paragraph{Phase 3: place.}
Walk children, assigning each one an $(x, y)$. The cursor advances
by the child's final main-axis size plus the gap; the cross-axis
position is determined by \code{align}. If \code{justify} is
non-trivial, the gap is enlarged or padding inserted at the ends.

Each child's final \code{computed} rectangle is written into the
\code{LayoutNode} struct, and the algorithm recurses into the
child to lay out \emph{its} children, with the child's resolved
size as the new available area.

\subsection{Wrapping}

When \code{wrap = Wrap} and the children don't fit on one line,
phase 2 partitions them into lines greedily: pack until the next
child would overflow, start a new line. Each line is then
independently laid out and placed at increasing cross-axis offsets.

For a wrapping column (rare), the lines are columns and the cross
axis is horizontal. Symmetric.

\section{Debugging layout: the four-step ritual}
\label{sec:layout:debugging}

When something is one cell off, two cells over, or in the wrong
place entirely, the cause is almost always one of four things.
Worth checking each in order before opening a debugger.

\paragraph{1. Check the available size.} The terminal's columns
and rows are passed in at the top of the layout. Print them. A
resize between frames will sometimes lag a frame, and you might be
laying out into the wrong rectangle.

\paragraph{2. Check padding and border on every ancestor.} A box
with padding 1 and a border eats 4 cells of width before its
children see anything. Inside a $24$-column terminal, after a
bordered-padded panel and another bordered-padded inner box,
children get 24 - 4 - 4 = 16 cells. The defect is rarely the layout;
it is usually an extra padding nobody remembered.

\paragraph{3. Print the layout tree.} A small helper:

\begin{lstlisting}
void dump(const LayoutNode& n, int depth = 0) {
    std::string indent(depth * 2, ' ');
    std::println("{}{}: ({},{}) {}x{} basis={} grow={}",
        indent, kind_name(n),
        n.computed.x, n.computed.y,
        n.computed.w, n.computed.h,
        n.style.basis, n.style.grow);
    for (int c : n.children) dump(nodes[c], depth + 1);
}
\end{lstlisting}

The answer is almost always visible in the first three lines of
output.

\paragraph{4. Use the DSL's debug border.} Wrap the suspicious
node in a border (\code{| border\_<Single>}); the corner glyphs
show you where its rectangle actually is. If the border is one
cell off, the layout is one cell off.

\begin{tryit}
Run any \maya{} sample and resize your terminal slowly. The
layout reflows on every \code{SIGWINCH}; you should see boxes
rearrange in cell steps, never half a cell. Try narrowing until
your status bar wraps; the wrap point is where total basis +
gaps exceeds the new available width.
\end{tryit}

\section{Layout recipes}
\label{sec:layout:recipes}

A handful of common shapes and how to spell them in the DSL.

\paragraph{Sidebar + content.}

\begin{lstlisting}
h(
    sidebar  | w_<24>,
    content  | grow_<1>
)
\end{lstlisting}

Sidebar fixed at 24; content takes everything else.

\paragraph{Header + body + footer.}

\begin{lstlisting}
v(
    header,
    body | grow_<1>,
    footer
)
\end{lstlisting}

Header and footer take their natural height; body fills the
leftover vertical space.

\paragraph{Three-pane equal split.}

\begin{lstlisting}
h(
    pane_a | grow_<1>,
    pane_b | grow_<1>,
    pane_c | grow_<1>
) | gap_<1>
\end{lstlisting}

Three equal-width panes with one-cell gaps. Each gets
$(W - 2) / 3$ cells.

\paragraph{Status bar on the right.}

\begin{lstlisting}
h(
    text(left),
    space,                  // spacer (grow=1, basis=0)
    text(right) | Dim
)
\end{lstlisting}

The spacer absorbs all leftover space, pushing the right text to
the edge.

\paragraph{Centred dialog.}

\begin{lstlisting}
h(
    space,
    v(
        space,
        dialog | w_<60>,
        space
    ),
    space
)
\end{lstlisting}

Four spacers form a cross-shaped frame around a 60-column dialog,
centring it horizontally and vertically.

\paragraph{A list with a fixed header.}

\begin{lstlisting}
v(
    header_row | Bold,
    sep,
    map(items, render_row) | grow_<1>
)
\end{lstlisting}

Header takes its row, separator one row, the body takes the rest.
When the list overflows, the \code{| grow\_<1>} on the mapped
content makes its container scrollable (Chapter~\ref{ch:widgets}
covers scroll surfaces).

\section{Performance notes}
\label{sec:layout:perf}

Layout is $O(n)$ in the number of nodes for an unwrapped tree.
With wrap, the worst case is $O(n^2)$ (every child triggers a
re-evaluation of remaining line capacity), but typical trees do
not wrap and the constant factor is small.

Numbers from the maya benchmark suite, on a typical desktop CPU:

\begin{itemize}[leftmargin=*]
  \item A 50-node tree (typical small app): under 5\,\textmu{}s.
  \item A 500-node tree (a long list of items): around
    30\,\textmu{}s.
  \item A 5000-node tree (pathological --- a maximalist
    dashboard): around 300\,\textmu{}s.
\end{itemize}

Layout is rarely the bottleneck. If it shows up in your profiler,
the usual cause is rebuilding too many DSL nodes per frame, not
the solver itself. Flatten where you can; cache \code{constexpr}
subtrees that do not change.

\section{Pitfalls and anti-patterns}
\label{sec:layout:pitfalls}

\begin{warning}
\textbf{Negative space from over-applied padding.} If parent and
child both apply padding, the visible insets compound. Decide which
layer owns the padding and keep it there.
\end{warning}

\begin{warning}
\textbf{A grow child inside a non-growing parent.} The parent sizes
itself to content, so there is no leftover space to distribute, and
the child does not grow. To get a growing layout, every ancestor up
to the root must be growing or explicitly sized.
\end{warning}

\begin{warning}
\textbf{Mixing \code{shrink=0} children with insufficient width.}
If every child refuses to shrink and they collectively exceed the
parent's width, content overflows. The renderer clips; the user
sees text cut off mid-glyph. Allow at least the variable-content
child to shrink.
\end{warning}

\begin{warning}
\textbf{Forgetting that text has intrinsic width.} A
\code{TextElement} in a row sizes to its content width by default.
Long strings push siblings off the edge. Wrap them in a fixed-width
box (\code{| w\_<30>}) or set their wrap mode to
\code{TextWrap::Truncate}.
\end{warning}

\begin{recap}
Layout decides where things go. \maya{} uses an integer-cell flexbox
solver over a flat \code{vector<LayoutNode>} addressed by index.
Three phases: measure, distribute, place. The same model handles
rows, columns, wrapping, alignment, justification, and nested boxes.
\end{recap}

\section*{Workshop}

\begin{enumerate}[leftmargin=*]
  \item Write a row that displays your name on the left and the date
    on the right, with the space between them growing to fill the
    terminal width.
  \item Build a 3-column dashboard: three boxes, each with
    \code{grow=1}, separated by 1-cell gaps. Resize your terminal
    --- watch the boxes split the new width.
  \item Read \file{maya/src/layout/yoga.cpp}. Identify the three
    phases (measure, distribute, place). Note where the available
    width is subdivided.
\end{enumerate}

\section{Layout, as a constraint problem}

A layout engine answers one question: given a tree of boxes with
properties, and a fixed-size terminal, where does each box go?
It sounds simple. It is not. The complication lives in
``properties'': intrinsic sizes, min/max constraints, grow/shrink
factors, padding, borders, gaps, wrapping. A real layout engine
must resolve all of them simultaneously.

The job has three classical solutions:

\begin{itemize}[leftmargin=*]
  \item \textbf{Tables} (HTML 4, ASCII art, dialog box managers).
    Cells in a fixed grid. Excellent for tabular data, terrible
    for everything else.
  \item \textbf{Constraints} (Cassowary, Auto Layout, Cocoa).
    The user expresses ``A.right = B.left + 8'' and the solver
    finds a satisfying assignment. Powerful, but opaque when it
    cannot find one --- and the solver is non-trivial.
  \item \textbf{Flow} (CSS, flexbox, grid). A small set of
    high-level intents (``these children should share the row
    equally'') that compile to a deterministic algorithm.
\end{itemize}

Flexbox is the flow solution. It is what every modern UI
framework uses (CSS, Yoga in React Native, Taffy in Rust UI
libraries) because it covers the practical cases with a small
set of mechanical rules.

\maya{}'s engine is a flexbox solver over integer cells. It is
smaller than CSS's flexbox (no fractional sizing, no baseline
alignment, no aspect ratios) but mechanically the same.

\subsection*{Why not the obvious algorithms?}

The naive ``put A next to B'' algorithm works until you have one
of these:

\begin{itemize}[leftmargin=*]
  \item A child that has no intrinsic size (a flex spacer).
  \item Two children that both want all the extra space.
  \item A row narrower than the sum of its children's preferred
    sizes.
  \item Wrapping.
  \item A box with padding plus a border plus content.
\end{itemize}

Any of those cases is enough to send a hand-rolled layout into
special-case soup. Flexbox is the algorithm that handles all of
them with one set of rules.

\begin{funfact}
The word ``flexbox'' is short for ``flexible box''. The CSS
spec uses ``flex container'' and ``flex item''; the shorthand
stuck. The spec was first drafted in 2009 by Tab Atkins at
Google and finalised in 2017. Yoga (Facebook's C++ flexbox)
predates CSS conformance and inspired \maya{}'s solver.
\end{funfact}

\section{The three phases, expanded}

\maya{}'s layout pass runs in three phases. Each phase produces
data the next phase consumes; there is no backtracking.

\subsection*{Phase 1: measure}

For each leaf node, compute its intrinsic size. A text node's
intrinsic width is the longest line; its intrinsic height is the
number of lines. A box's intrinsic size is computed recursively:

\begin{lstlisting}
Size measure(const Element& e) {
    return std::visit(overload{
        [](const TextElement& t) {
            auto [w, h] = grapheme_extent(t.text);
            return Size{w, h};
        },
        [](const BoxElement& b) {
            int w = 0, h = 0;
            for (auto& c : b.children) {
                Size cs = measure(c);
                if (b.layout.direction == Direction::Row) {
                    w += cs.w;
                    h = std::max(h, cs.h);
                } else {
                    h += cs.h;
                    w = std::max(w, cs.w);
                }
            }
            // Add padding and border to the box's own size.
            w += b.padding.horizontal() + (b.border ? 2 : 0);
            h += b.padding.vertical()   + (b.border ? 2 : 0);
            return Size{w, h};
        },
        [](const ElementList& l) {
            // Spliced into parent; measure is sum.
            int w = 0, h = 0;
            for (auto& c : l.items) {
                Size cs = measure(c);
                w = std::max(w, cs.w); h += cs.h;
            }
            return Size{w, h};
        }
    }, e.v);
}
\end{lstlisting}

The phase is bottom-up: leaves first, then their parents, then
the root. After this phase every node knows ``how big would I be
if nobody constrained me?''.

\subsection*{Phase 2: distribute}

Given a parent's actual size (constrained by the terminal or its
own parent), how much does each child get?

For a row of children with grow factors $g_1, g_2, \ldots, g_n$
and intrinsic widths $w_1, w_2, \ldots, w_n$, and an available
width $W$:

\[
  \text{slack} = W - \sum w_i
\]

If slack is positive (the row has extra room):

\[
  w_i' = w_i + g_i \cdot \frac{\text{slack}}{\sum g_j}
\]

If slack is negative (the row is over-budget) and the children
have \emph{shrink} factors $s_1, \ldots, s_n$:

\[
  w_i' = w_i - s_i \cdot \frac{|\text{slack}|}{\sum s_j}
\]

The rounded integer cells must sum to exactly $W$, so the
remainder is distributed cell by cell to the first children
that asked for grow.

\subsection*{Phase 3: place}

Now each child has its size. Walk the row again, placing each
child at the running $x$ position. After placing, recurse into
the child to lay out \emph{its} children.

\begin{lstlisting}
void place(const Element& e, Rect rect) {
    std::visit(overload{
        [&](const TextElement& t) {
            // Text occupies its assigned rect.
            register_text(t, rect);
        },
        [&](const BoxElement& b) {
            Rect inner = shrink(rect, b.padding, b.border);
            int pos = 0;
            for (auto& [child, size] : b.children_with_sizes) {
                Rect child_rect = (b.layout.direction == Direction::Row)
                    ? Rect{inner.x + pos, inner.y, size.w, inner.h}
                    : Rect{inner.x, inner.y + pos, inner.w, size.h};
                place(child, child_rect);
                pos += (b.layout.direction == Direction::Row) ? size.w : size.h;
            }
            register_box(b, rect);
        },
        [&](const ElementList& l) {
            int pos = 0;
            for (auto& c : l.items) {
                Rect cr{rect.x, rect.y + pos, rect.w, /* size from phase 2 */};
                place(c, cr);
                pos += cr.h;
            }
        }
    }, e.v);
}
\end{lstlisting}

After this phase every node has a final \code{Rect}: $x$, $y$,
width, height. The canvas-drawing phase (Chapter~\ref{ch:canvas})
consumes those rects.

\section{Every flex property, mechanically}

We will now walk every property on \code{FlexStyle}, with a
worked example for each.

\subsection*{Direction}

Row or Column. Determines whether children stack horizontally or
vertically.

\begin{lstlisting}
FlexStyle{.direction = Direction::Row}     // children stack left-to-right
FlexStyle{.direction = Direction::Column}  // top-to-bottom
\end{lstlisting}

The DSL spells these as \code{h(\ldots)} and \code{v(\ldots)}.
Reverse variants (\code{RowReverse}, \code{ColumnReverse}) flip
the order without reflowing.

\subsection*{Justify}

Where children sit along the main axis. Six values:

\begin{itemize}[leftmargin=*]
  \item \code{Start}: children at the start; trailing space.
  \item \code{End}: children at the end; leading space.
  \item \code{Center}: children centred; equal space on both sides.
  \item \code{SpaceBetween}: first/last at the edges; equal space
    between siblings.
  \item \code{SpaceAround}: equal space around each child
    (half-gap at the edges).
  \item \code{SpaceEvenly}: equal gaps everywhere, including
    edges.
\end{itemize}

A worked example. Three children of width 3, 4, 5 in a row of
width 20 (slack = 8):

\begin{terminalbox}
Start        : AAABBBBCCCCC........\\
End          : ........AAABBBBCCCCC\\
Center       : ....AAABBBBCCCCC....\\
SpaceBetween : AAA....BBBB....CCCCC\\
SpaceAround  : ..AAA..BBBB..CCCCC..\\
SpaceEvenly  : ..AAA..BBBB..CCCCC..
\end{terminalbox}

Notice SpaceAround and SpaceEvenly look identical here because
the gaps happen to divide evenly. With slightly different numbers
they diverge (SpaceEvenly always equalises all gaps; SpaceAround
halves the edge gaps).

\subsection*{Align}

Where children sit along the cross axis. Five values:

\begin{itemize}[leftmargin=*]
  \item \code{Start}: top (for rows) or left (for columns).
  \item \code{End}: bottom or right.
  \item \code{Center}: centred along the cross axis.
  \item \code{Stretch}: child grows to fill the cross axis
    (default for boxes without an intrinsic cross size).
  \item \code{Baseline}: align by the baseline of the first
    text line (not yet implemented in \maya{}; degrades to
    Start).
\end{itemize}

Two children of heights 2 and 4 in a row of height 5:

\begin{terminalbox}
Start   : AB  | AB    Center: AB    Stretch: AAB\\
          A B | A B           AB             A B\\
            B |   B          A B             A B\\
              |              A B               B\\
              |                B               B
\end{terminalbox}

\subsection*{Grow}

A non-negative number. Children with positive grow share extra
space proportionally. \code{grow=0} means ``do not grow''.
\code{grow=1} and \code{grow=2} split slack 1:2.

Worked example. Two children with intrinsic widths 5 and 5,
grow 1 and 3, in a row of width 24:

\begin{itemize}[leftmargin=*]
  \item Slack: $24 - (5+5) = 14$.
  \item Grow sum: $1 + 3 = 4$.
  \item Child A: $5 + 14 \cdot 1/4 = 5 + 3.5 \to 5 + 3 = 8$
    (rounded down).
  \item Child B: $5 + 14 \cdot 3/4 = 5 + 10.5 \to 5 + 11 = 16$
    (rounded up to absorb the remainder).
  \item Total: $8 + 16 = 24$. \checkmark
\end{itemize}

If $5 + 3 = 8$ and $5 + 10 = 15$, the sum would be 23 ---
 short by one cell. The remainder distribution awards the
leftover cell to the first child that grew, which is why
rounding sometimes favours one side.

\subsection*{Shrink}

The mirror of grow. When the row is over-budget, children with
shrink absorb the deficit proportionally. \code{shrink=0} means
``never shrink me; overflow the row instead''.

Default: \code{shrink=1} for all children. This means rows
naturally accommodate; set \code{shrink=0} on a fixed-width
element to make it inviolate.

\subsection*{Basis}

The starting size before grow/shrink. If unset, basis = intrinsic
size. If set, basis overrides the intrinsic measurement.

\begin{lstlisting}
FlexStyle{.basis = 10, .grow = 1}  // start at 10, grow from there
\end{lstlisting}

Useful when you want a minimum reservation that may grow but not
shrink below.

\subsection*{Wrap}

What to do if the row's children exceed the available width:

\begin{itemize}[leftmargin=*]
  \item \code{NoWrap} (default): try to shrink, otherwise overflow.
  \item \code{Wrap}: spill onto subsequent rows.
  \item \code{WrapReverse}: spill, but stack rows from bottom up.
\end{itemize}

Wrap is what makes a tag-cloud layout possible. The solver places
as many children as fit on the first line, then starts a new line.

\subsection*{Gap}

Fixed cell-spacing between children. Subtracted from available
size before distribution:

\begin{lstlisting}
FlexStyle{.direction = Direction::Row, .gap = 2}
\end{lstlisting}

With $n$ children in a row of width $W$ and gap $g$, the
available for children is $W - (n-1) \cdot g$.

\subsection*{Padding}

Per-side integer cells, inside the box. The content area is the
box area shrunk by padding.

\code{Padding\{.top=1, .right=2, .bottom=1, .left=2\}}, or the
shorthand \code{Padding::all(1)}, or \code{pad<1, 2>} from the
DSL.

\subsection*{Border}

Borders take one cell per side. They are not padding but they
are also subtracted from the content area.

\section{The integer remainder strategy}

The distribution formula gives fractional cell counts. Cells are
integers. We must round.

The naive strategy (round each independently) often misses the
target by one cell. We could be one cell short or one over. The
strategy in \maya{}:

\begin{enumerate}[leftmargin=*]
  \item For each child, compute floor of the grow-distributed
    size. Sum the floors.
  \item Compute the remainder: $r = W - \sum \lfloor w_i' \rfloor$.
  \item Award one extra cell each, in order, to the first $r$
    children that had a fractional part. Children with the
    largest fractional part go first (Hamilton's method, the
    same algorithm US Congress used to apportion seats).
\end{enumerate}

This guarantees:

\begin{itemize}[leftmargin=*]
  \item Sums work out: $\sum w_i' = W$ exactly.
  \item Stable: rounding favours the children who asked for
    fractional cells, not arbitrarily.
  \item Fair: no single child can be more than one cell off
    from its ideal share.
\end{itemize}

\begin{idea}
This is the same problem (and the same solution) that political
science studies under ``apportionment''. The Alabama paradox
(adding a state can decrease another state's seats) does not
apply here because the total $W$ does not change between
rounds, but the spirit is identical.
\end{idea}

\section{Four reasons cells are integers, not floats}

A recurring question: why not use floats? The reasons add up.

\begin{enumerate}[leftmargin=*]
  \item \textbf{Terminals are integer grids.} You cannot draw at
    x=5.5. The renderer would have to round at the end anyway,
    and floating-point round can drift.
  \item \textbf{Determinism.} Integer arithmetic produces
    bit-identical results across machines. Float arithmetic
    does not (different FPU modes, different compiler flags).
    For replay and tests, determinism matters.
  \item \textbf{Performance.} Integer add/multiply is the same
    or faster than float; integer comparisons are unambiguous.
    No NaN, no infinity, no subnormals.
  \item \textbf{Avoiding fractional-cell bugs.} If your widget
    expects ``width 30'' and gets ``29.999998'', it might emit
    one fewer character than expected. With integers, the
    contract is precise.
\end{enumerate}

\begin{warning}
If you ever see \maya{}'s layout produce a non-integer size in
a debug dump, that is a bug; please file it. The whole pipeline
is integer-typed.
\end{warning}

\section{Worked traces: three real layouts}

\subsection*{Trace 1: a status bar}

Source:

\begin{lstlisting}
h(text("NORMAL") | grow<0>,
  text("~/file.txt") | grow<1>,
  text("Ln 12  Col 3") | grow<0>)
  | width<60>
\end{lstlisting}

Measure phase:

\begin{itemize}[leftmargin=*]
  \item Child A ("NORMAL"): intrinsic width 6.
  \item Child B ("~/file.txt"): intrinsic width 11.
  \item Child C ("Ln 12  Col 3"): intrinsic width 12.
  \item Row intrinsic: $6 + 11 + 12 = 29$.
\end{itemize}

Distribute (W = 60, slack = $60 - 29 = 31$, only B has grow=1):

\begin{itemize}[leftmargin=*]
  \item A: 6 (no grow).
  \item B: $11 + 31 = 42$.
  \item C: 12 (no grow).
  \item Sum: $6 + 42 + 12 = 60$. \checkmark
\end{itemize}

Place:

\begin{terminalbox}
NORMAL ~/file.txt                              Ln 12  Col 3
\end{terminalbox}

\subsection*{Trace 2: a dashboard}

Source:

\begin{lstlisting}
h(box(text("left"))   | grow<1>,
  box(text("middle")) | grow<2>,
  box(text("right"))  | grow<1>)
  | width<40>
\end{lstlisting}

Measure: each box has intrinsic 4 + 2 (padding) + 2 (border) =
approx 8. Row intrinsic = 24. Slack = 16. Grow sum = 4. Per
grow-unit: 4.

Distribute:

\begin{itemize}[leftmargin=*]
  \item Left:   $8 + 4 = 12$.
  \item Middle: $8 + 8 = 16$.
  \item Right:  $8 + 4 = 12$.
  \item Sum: 40. \checkmark
\end{itemize}

The middle column is twice as wide as the side columns, which is
what the user asked for.

\subsection*{Trace 3: an over-budget row}

Source:

\begin{lstlisting}
h(text("hello, world")    | shrink<1>,
  text("a very long line") | shrink<2>)
  | width<20>
\end{lstlisting}

Measure: 12 + 15 = 27. Available 20. Deficit = 7. Shrink sum = 3.
Per shrink-unit: $7 / 3 \approx 2.33$.

Distribute:

\begin{itemize}[leftmargin=*]
  \item Child 1: $12 - 1 \cdot 2.33 \approx 10$ (truncated
    via the wrap policy or ellipsis).
  \item Child 2: $15 - 2 \cdot 2.33 \approx 10$.
\end{itemize}

The text is clipped to fit. If a child had \code{shrink=0}, it
would keep its intrinsic size and the other child would absorb
the entire deficit.

\section{A debugging ritual}

When a layout looks wrong, four steps:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Dump the tree.} Use the \code{dump} helper from
    Chapter~\ref{ch:elements}. Confirm the structure matches
    what you intended.
  \item \textbf{Print the rects.} \maya{} has a
    \code{dump\_layout(\ldots)} helper that prints each node's
    final rect. Look for surprises (a child wider than its
    parent, a rect with negative width).
  \item \textbf{Walk the row sums.} For each row, the sum of
    children's widths plus gaps plus padding plus border must
    equal the row's width. If it does not, a grow/shrink
    setting is wrong.
  \item \textbf{Resize the terminal.} A layout that looks fine
    at 80 columns but breaks at 100 is a grow-factor mistake.
    Try several widths.
\end{enumerate}

A \code{dump\_layout} sample:

\begin{lstlisting}
box 60x3 at (0,0)
  text "NORMAL"        6x1 at (0,1)
  text "~/file.txt"   42x1 at (6,1)
  text "Ln 12 Col 3"  12x1 at (48,1)
\end{lstlisting}

Note that the box is 3 high but the text rows are at y=1 ---
the border takes y=0. If the y were 0 with no border, that would
be a clue.

\section{Layout recipes}

A gallery of common shapes and how they decompose to flex.

\subsection*{Centred dialog}

\begin{lstlisting}
v(box(content) | align<Center> | grow<1>);
\end{lstlisting}

The outer column has one child that grows. Setting
\code{align=Center} on the column centres the child on the
cross axis (horizontally, for a column).

\subsection*{Sidebar + main}

\begin{lstlisting}
h(sidebar | width<24>,
  main    | grow<1>);
\end{lstlisting}

Fixed-width sidebar, flexible main area. No special property
required --- grow does the work.

\subsection*{Header + body + footer}

\begin{lstlisting}
v(header | height<1>,
  body   | grow<1>,
  footer | height<1>);
\end{lstlisting}

Fixed-height header and footer; body absorbs the rest.

\subsection*{Tag cloud (wrapping row)}

\begin{lstlisting}
h(tags...) | wrap<Wrap> | gap<1>;
\end{lstlisting}

Tags flow left to right; when a row fills, the next tag starts
on a new line.

\subsection*{Two-column form}

\begin{lstlisting}
v(
  h(text("Name:")  | width<10>, text(name_value)),
  h(text("Email:") | width<10>, text(email_value)),
  h(text("Notes:") | width<10>, text(notes_value))
);
\end{lstlisting}

Labels share a fixed left column; values fill the rest. Notice
the DSL has no ``column'' primitive --- the alignment falls out
of identical \code{width<10>} on each label.

\subsection*{Centred status with action buttons}

\begin{lstlisting}
h(
  text("Saved") | align<Center> | grow<1>,
  text("[OK]")  | grow<0>
);
\end{lstlisting}

Left text centred in the flexible region, right button fixed.

\section{Layout performance}

The whole pass is $O(n)$ in tree size. For a typical UI of a few
hundred nodes, expect:

\begin{itemize}[leftmargin=*]
  \item Measure phase: 10--30 microseconds.
  \item Distribute phase: 5--15 microseconds.
  \item Place phase: 5--10 microseconds.
  \item Total: well under 60 microseconds on a modern CPU.
\end{itemize}

The heavy work is mostly memory: the layout pass allocates one
flat \code{vector<LayoutNode>} keyed by tree-walk index. The
vector is reused frame to frame (capacity persists, only size is
reset), so steady-state allocation is zero.

\begin{aside}
Flexbox in browsers is slower because CSS adds many properties
(min/max sizing, baselines, aspect ratios, fractional units,
writing modes). The simpler integer-cell model in \maya{} cuts
perhaps 80\% of the work.
\end{aside}

\section{Pitfalls beyond the obvious}

\begin{warning}
\textbf{Forgetting that a child without a size hint has zero
intrinsic.} An empty box with no content and no width has
intrinsic (0, 0). If you stick it in a row with other children
and expect it to take space, it will not. Add \code{width<n>}
or \code{grow<n>}.
\end{warning}

\begin{warning}
\textbf{Border + padding + content = total.} If you want a box
that fits in a row of width 20 with a border, the content must
fit in $20 - 2 = 18$. Forgetting the border eats two cells you
will miss.
\end{warning}

\begin{warning}
\textbf{Mixing grow with fixed width.} \code{width<10>} sets the
intrinsic; \code{grow<1>} adds slack on top. Using both gives a
child that starts at 10 and grows. If you wanted exactly 10,
omit grow.
\end{warning}

\begin{warning}
\textbf{Negative cells from over-shrinking.} If shrink lets a
child reach width 0, then 1, then -1, things go sideways.
\maya{} clamps at zero, but the result is invisible. Set
\code{shrink=0} on children that must keep their minimum size.
\end{warning}

\begin{warning}
\textbf{Wrap interacting with grow.} On a wrapping row, grow
factors operate \emph{within each line}, not across the whole
row. The first line and the second line both distribute their
own slack independently.
\end{warning}

\section{The bigger picture}

Layout is the bridge from the element tree to the canvas. Once
every node has a \code{Rect}, the canvas-paint phase
(Chapter~\ref{ch:canvas}) walks the tree and writes packed cells
at the assigned coordinates.

The layout pass is the most complex single piece of \maya{}'s
render pipeline, but it is still just $O(n)$ in tree size and
stays under a millisecond for any reasonable UI. Understanding
it pays off when you need a non-obvious shape or when you are
debugging why a child appears in the wrong place.

\begin{recap}
Flexbox in three phases: measure (bottom-up intrinsic sizes),
distribute (apportion slack/deficit by grow/shrink), place (assign
rects). Integer cells throughout, Hamilton apportionment for
rounding, $O(n)$ in tree size. Eight properties on
\code{FlexStyle} cover every shape; six recipes cover almost
every layout you will ever want.
\end{recap}
